---
title: "Main figure generation"
output:
  html_document:
    toc: true
params:
  execute: false
---

This notebook follows the final manuscript order for Visium Figures 2–4. Statistical calculations are performed in `Visium_analyses.Rmd`; this notebook reads the resulting objects and tables and generates the figure panels.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries, cache = FALSE}
library(Seurat)
library(tidyverse)
library(qs)
library(here)
library(patchwork)
library(dittoSeq)
library(ggsci)
source(here("R", "plotting.R"))
source(here("R", "scoring.R"))

combined_final <- qread(here("data", "derived", "visium_scored.qs"))
if (!"SSIE" %in% colnames(combined_final@meta.data)) {
  ssie_score <- calculate_ssie_cooccurrence(combined_final@meta.data)
  combined_final$SSIE <- ssie_score[colnames(combined_final)]
}
epithelial <- qread(here("data", "derived", "epithelial_diffusion_map.qs"))
cluster_composition_tests <- readRDS(
  here("results", "tables", "cluster_composition_tests.rds")
)
moran_composition_tests <- readRDS(
  here("results", "tables", "local_moran_composition_tests.rds")
)

# The diffusion map is calculated before Cell2location and local Moran's I.
# Attach any metadata added later by barcode (including both groups of results).
diffusion_metadata_columns <- setdiff(
  colnames(combined_final@meta.data),
  colnames(epithelial@meta.data)
)
if (length(diffusion_metadata_columns)) {
  stopifnot(all(colnames(epithelial) %in% rownames(combined_final@meta.data)))
  epithelial <- AddMetaData(
    epithelial,
    combined_final@meta.data[
      colnames(epithelial),
      diffusion_metadata_columns,
      drop = FALSE
    ]
  )
}
figure_dir <- here("results", "figures", "main")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

atlas_examples <- c("R3_large_4", "Pilot_small_1")
advanced_examples <- c("R4_large_1", "R4_large_2")
main_spatial_example <- "R4_large_2"
moran_levels <- c( "High-High","High-Low", "Low-High", "Low-Low")

p_to_stars <- function(p) {
  case_when(
    is.na(p) ~ "",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ ""
  )
}

plot_cluster_composition <- function(
    metadata,
    condition_col,
    condition_keep,
    posthoc = NULL
) {
  condition_levels <- recode(
    condition_keep,
    Small = "Non-Adv",
    Large = "Advanced",
    `Low-grade` = "Adenoma"
  )
  plot_data <- metadata |>
    filter(.data[[condition_col]] %in% condition_keep) |>
    transmute(
      condition = recode(
        as.character(.data[[condition_col]]),
        Small = "Non-Adv",
        Large = "Advanced",
        `Low-grade` = "Adenoma"
      ),
      condition = factor(condition, levels = condition_levels),
      cluster = factor(as.character(combined_cluster), levels = as.character(0:8))
    ) |>
    count(condition, cluster, .drop = FALSE) |>
    group_by(condition) |>
    mutate(proportion = n / sum(n)) |>
    ungroup()

  plot <- ggplot(plot_data, aes(condition, proportion, fill = cluster)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = cluster_colors, drop = FALSE) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    theme_classic() +
    labs(x = NULL, y = "Relative abundance", fill = "Cluster")

  if (!is.null(posthoc)) {
    category_col <- colnames(posthoc)[[1]]
    annotation_data <- plot_data |>
      filter(condition == last(condition_levels)) |>
      arrange(cluster) |>
      mutate(y = cumsum(proportion) - proportion / 2) |>
      left_join(
        posthoc |>
          transmute(
            cluster = as.character(.data[[category_col]]),
            label = p_to_stars(p_adj)
          ),
        by = "cluster"
      ) |>
      filter(label != "")
    plot <- plot + geom_text(
      data = annotation_data,
      aes(x = condition, y = y, label = label),
      inherit.aes = FALSE,
      size = 3
    )
  }
  plot
}

plot_moran_composition <- function(
    metadata,
    class_col,
    condition_col,
    condition_keep,
    posthoc = NULL
) {
  condition_levels <- recode(
    condition_keep,
    Small = "Non-Adv",
    Large = "Advanced",
    `Low-grade` = "Adenoma"
  )
  plot_data <- metadata |>
    filter(
      .data[[condition_col]] %in% condition_keep,
      !is.na(.data[[class_col]]),
      .data[[class_col]] != "Not sig"
    ) |>
    transmute(
      condition = recode(
        as.character(.data[[condition_col]]),
        Small = "Non-Adv",
        Large = "Advanced",
        `Low-grade` = "Adenoma"
      ),
      condition = factor(condition, levels = condition_levels),
      moran_class = factor(as.character(.data[[class_col]]), levels = moran_levels)
    ) |>
    count(condition, moran_class, .drop = FALSE) |>
    group_by(condition) |>
    mutate(proportion = n / sum(n)) |>
    ungroup()

  plot <- ggplot(plot_data, aes(condition, proportion, fill = moran_class)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = moran_colors, drop = FALSE) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    theme_classic() +
    labs(x = NULL, y = "Relative abundance", fill = NULL)

  if (!is.null(posthoc)) {
    category_col <- colnames(posthoc)[[1]]
    annotation_data <- plot_data |>
      filter(condition == last(condition_levels)) |>
      arrange(moran_class) |>
      mutate(y = cumsum(proportion) - proportion / 2) |>
      left_join(
        posthoc |>
          transmute(
            moran_class = as.character(.data[[category_col]]),
            label = p_to_stars(p_adj)
          ),
        by = "moran_class"
      ) |>
      filter(label != "")
    plot <- plot + geom_text(
      data = annotation_data,
      aes(x = condition, y = y, label = label),
      inherit.aes = FALSE,
      color = "white",
      size = 3
    )
  }
  plot
}
```

# Figure 2: Integrated multimodal annotation

## Figure 2A: integrated clusters

```{r figure-2a-clusters}
p2a_umap <- DimPlot(
  combined_final,
  reduction = "wnn.umap",
  group.by = "combined_cluster",
  cols = cluster_colors
) + theme_void()

p2a_spatial <- plot_spatial_samples(
  combined_final,
  group_by = "combined_cluster",
  samples = atlas_examples,
  colors = cluster_colors,
  ncol = 1
)

p2a <- p2a_umap | p2a_spatial
p2a
save_plot(p2a, file.path(figure_dir, "Figure2A_clusters.pdf"), 9, 6)
```

## Figure 2B: curated RNA markers

RNA markers are grouped by their annotated cell class.

```{r figure-2b-markers}
rna_markers <- c(
  "LGR5", "ASCL2", "OLFM4", "AXIN2", "EPHB2", "ALDH1A1", "ALCAM",
  "MKI67", "TOP2A", "PCNA",
  "CDKN1A", "CDKN2B", "TP53", "GLB1", "CXCL1",
  "SLC26A3", "CA2", "CLCA1", "TFF3", "CEACAM7",
  "COL3A1", "DES", "GREM1",
  "TRAC", "MS4A1", "PTPRC"
)
rna_marker_classes <- tibble(
  features.plot = rna_markers,
  class = c(
    rep("Stemness", 7),
    rep("Proliferation", 3),
    rep("Senescence / stress", 5),
    rep("Normal / differentiated epithelium", 5),
    rep("Stromal", 3),
    rep("Immune", 3)
  )
)

DefaultAssay(combined_final) <- "Spatial"
p2b <- DotPlot_by_class_horizontal(
  object = combined_final,
  features = rna_markers,
  feature_class = rna_marker_classes,
  col.low = "#053061",
  col.mid = "#F5F6F6",
  col.high = "#67001F"
)
p2b
save_plot(p2b, file.path(figure_dir, "Figure2B_RNA_markers.pdf"), 8, 6)
```

## Figure 2C: pathology annotations

```{r figure-2c-pathology}
p2c_umap <- DimPlot(
  combined_final,
  reduction = "wnn.umap",
  group.by = "Adenoma",
  cols = region_colors
) + theme_void()

p2c_spatial <- plot_spatial_samples(
  combined_final,
  group_by = "Adenoma",
  samples = atlas_examples,
  colors = region_colors,
  ncol = 1
)

p2c <- p2c_umap | p2c_spatial
p2c
save_plot(p2c, file.path(figure_dir, "Figure2C_pathology.pdf"), 9, 6)
```

## Figure 2D: cluster composition

The plotted proportions are descriptive spot compositions. The sample-aware tests are saved by the `sample-level-statistics` chunk of `Visium_analyses.Rmd`.

```{r figure-2d-cluster-composition}
cluster_composition_tests = read_rds(here("results","tables","cluster_composition_tests.rds"))
p2d_size <- plot_cluster_composition(
  combined_final@meta.data,
  condition_col = "LargeSmall",
  condition_keep = c("Small", "Large"),
  posthoc = cluster_composition_tests$largesmall$posthoc
)
p2d_region <- plot_cluster_composition(
  combined_final@meta.data,
  condition_col = "Adenoma",
  condition_keep = c("Normal", "Low-grade"),
  posthoc = cluster_composition_tests$adenoma$posthoc
)
p2d <- p2d_size | p2d_region
p2d
save_plot(p2d, file.path(figure_dir, "Figure2D_cluster_composition.pdf"), 8, 4)
```

## Figure 2E: spatial pseudotime

```{r figure-2e-spatial-pseudotime}
p2e_umap <- SCP::FeatureDimPlot(
  combined_final,
  reduction = "wnn.umap",
  features = "dpt_pseudotime"
) + theme_void()
p2e_spatial <- SpatialFeaturePlot(
  combined_final,
  features = "dpt_pseudotime",
  images = atlas_examples,
  ncol = 1,
  keep.scale = "all"
)
p2e <- p2e_umap | p2e_spatial
p2e
save_plot(p2e, file.path(figure_dir, "Figure2E_spatial_pseudotime.pdf"), 9, 6)
```

## Figure 2F: pseudotime by pathology and advancement

```{r figure-2f-pseudotime-pathology}
p2f <- dittoSeq::dittoRidgePlot(
  subset(combined_final, subset = Adenoma != "Stroma"),
  var = "dpt_pseudotime",
  group.by = "Adenoma",
  split.by = "LargeSmall",
  y.breaks = c(0, 0.5, 1)
)
p2f
save_plot(p2f, file.path(figure_dir, "Figure2F_pseudotime_pathology.pdf"), 5, 4)
```

## Figure 2G: pseudotime by cluster and advancement

```{r figure-2g-pseudotime-clusters}
p2g <- dittoSeq::dittoRidgePlot(
  combined_final,
  var = "dpt_pseudotime",
  group.by = "combined_cluster",
  split.by = "LargeSmall",
  color.panel = unname(cluster_colors),
  y.breaks = c(0, 0.5, 1)
)
p2g
save_plot(p2g, file.path(figure_dir, "Figure2G_pseudotime_clusters.pdf"), 5, 4)
```

## Figure 2H: diffusion map

The diffusion-map object displays diffusion components 1 and 3.

```{r figure-2h-diffusion-clusters}
p2h <- DimPlot(
  epithelial,
  reduction = "diffmap",
  group.by = "combined_cluster",
  cols = cluster_colors
) + theme_void()
p2h
save_plot(p2h, file.path(figure_dir, "Figure2H_diffusion_clusters.pdf"), 5, 5)
```

## Figure 2I: diffusion-map scores

```{r figure-2i-diffusion-scores}
p2i <- SCP::FeatureDimPlot(
  epithelial,
  reduction = "diffmap",
  features = c(
    "dpt_pseudotime", "Senepy", "Stem_Signature",
    "WNT_BETA_CATENIN_SIGNALING", "P53_PATHWAY", "IEX"
  ),
  ncol = 2,
  keep_scale = "feature",
  show_stat = F,theme_use = SCP::theme_blank
)
p2i
save_plot(p2i, file.path(figure_dir, "Figure2I_diffusion_scores.pdf"), 8, 10)
```

# Figure 3: Senescence, stemness, and spatial heterogeneity

## Figure 3A: senescence and stemness UMAPs

```{r figure-3a-senescence-stemness}
p3a <- SCP::FeatureDimPlot(
  combined_final,
  reduction = "wnn.umap",
  features = c("Senepy", "Stem_Signature"),
  ncol = 1,
  keep_scale = "feature",
  show_stat = F,theme_use = SCP::theme_blank
) 
p3a
save_plot(p3a, file.path(figure_dir, "Figure3A_senescence_stemness.pdf"), 5, 8)
```

## Figure 3B: distance from the luminal surface

Stemness and senescence scores are summarized across distance from the annotated luminal surface.

```{r figure-3b-surface-distance}
stem_distance <- read_csv(
  here("results", "tables", "surface_distance_stemness.csv"),
  show_col_types = FALSE
)
senescence_distance <- read_csv(
  here("results", "tables", "surface_distance_senescence.csv"),
  show_col_types = FALSE
)
distance_colors <- c(
  Normal = unname(region_colors[["Normal"]]),
  Adenoma = unname(region_colors[["Low-grade"]])
)

surface_score_plot <- function(data, y_label) {
  ggplot(data, aes(distance_bin_um, mean_score, color = Adenoma)) +
    geom_point(aes(size = n_spots), alpha = 0.8) +
    geom_smooth(
      method = "gam",
      formula = y ~ s(x, bs = "cs"),
      se = TRUE,
      linewidth = 1
    ) +
    scale_color_manual(values = distance_colors) +
    coord_cartesian(xlim = c(0, 825)) +
    ggpubr::theme_pubr() +
    labs(
      x = "Distance from tract surface (micrometers)",
      y = y_label,
      size = "N spots",
      color = "Region"
    )
}

p3b_senescence <- surface_score_plot(
  senescence_distance,
  "Mean senescence score per 100-micrometer bin"
)
p3b_stemness <- surface_score_plot(
  stem_distance,
  "Mean stemness score per 100-micrometer bin"
)
p3b <- p3b_senescence / p3b_stemness
p3b
save_plot(p3b, file.path(figure_dir, "Figure3B_surface_distance.pdf"), 5, 8)
```

## Figure 3C: representative spatial senescence–stemness maps

```{r figure-3c-spatial-senescence-stemness}
p3c_pathology <- SpatialDimPlot(
  combined_final,
  group.by = "Adenoma",
  images = main_spatial_example,
  cols = region_colors
)
p3c_he <- SpatialDimPlot(
  combined_final,
  group.by = "Adenoma",
  images = main_spatial_example,
  cols = region_colors,
  alpha = 0
) + NoLegend()
p3c_scores <- SpatialFeaturePlot(
  combined_final,
  features = c("Senepy", "Stem_Signature"),
  images = main_spatial_example,
  ncol = 1,
  keep.scale = "feature"
)
p3c_moran <- SpatialDimPlot(
  combined_final,
  group.by = "Cluster_Moran_Senepy_Stem_Signature",
  images = main_spatial_example,
  cols = moran_colors
)
p3c_clusters <- SpatialDimPlot(
  combined_final,
  group.by = "combined_cluster",
  images = main_spatial_example,
  cols = cluster_colors
)
p3c <- wrap_plots(
  p3c_pathology, p3c_scores[[1]], p3c_moran,
  p3c_he, p3c_scores[[2]], p3c_clusters,
  ncol = 3
)
p3c
save_plot(p3c, file.path(figure_dir, "Figure3C_spatial_senescence_stemness.pdf"), 11, 7)
```

## Figure 3D: senescence–stemness Moran classes

```{r figure-3d-senescence-stemness-moran, fig.width=12,fig.height=5}
moran_composition_tests = read_rds(here("results","tables","local_moran_composition_tests.rds"))
p3d_size <- plot_moran_composition(
  filter(combined_final@meta.data, Adenoma == "Low-grade"),
  class_col = "Cluster_Moran_Senepy_Stem_Signature",
  condition_col = "LargeSmall",
  condition_keep = c("Small", "Large"),
  posthoc = moran_composition_tests$largesmall[[
    "Cluster_Moran_Senepy_Stem_Signature"
  ]]$posthoc
)
p3d_region <- plot_moran_composition(
  combined_final@meta.data,
  class_col = "Cluster_Moran_Senepy_Stem_Signature",
  condition_col = "Adenoma",
  condition_keep = c("Normal", "Low-grade"),
  posthoc = moran_composition_tests$adenoma[[
    "Cluster_Moran_Senepy_Stem_Signature"
  ]]$posthoc
)
p3d_diffusion <- DimPlot(
  epithelial,
  reduction = "diffmap",
  group.by = "Cluster_Moran_Senepy_Stem_Signature",
  cols = moran_colors
) + theme_void()
p3d <- p3d_size | p3d_region | p3d_diffusion
p3d
save_plot(p3d, file.path(figure_dir, "Figure3D_senescence_stemness_moran.pdf"), 11, 4)
```

## Figure 3E: three-way Cell2location abundance

```{r figure-3e-three-way-cell2location, fig.width=8,fig.height=5}
cell2location_features <- c("Dysplastic_Epi", "CD8_T_cells", "Normal_Epi")
p3e_umap <- three_way_reduction_plot(
  combined_final,
  features = cell2location_features,
  reduction = "wnn.umap"
)
p3e_spatial <- three_way_spatial_plot(
  combined_final,
  features = cell2location_features,
  images = main_spatial_example
)
p3e <- p3e_umap | p3e_spatial
p3e
save_plot(p3e, file.path(figure_dir, "Figure3E_three_way_cell2location.pdf"), 10, 5)
```

## Figure 3F: dysplastic epithelial subtypes in diffusion space

```{r figure-3f-dysplastic-subtypes-diffusion}
dysplastic_features <- c(
  "Dysplastic_Senescent_like",
  "Dysplastic_Stem_like",
  "Dysplastic_Tuft_like",
  "Dysplastic_Proliferating"
)
dysplastic_feature_labels <- c(
  "LGALS3+ Dysplastic cell",
  "LGR5+ Dysplastic cell",
  "AVIL+ Dysplastic cell",
  "MKI67+ Dysplastic cell"
)
p3f_plots <- SCP::FeatureDimPlot(
  epithelial,
  reduction = "diffmap",
  features = dysplastic_features,
  combine = FALSE,
  keep_scale = "feature",show_stat = F,theme_use = SCP::theme_blank
) |>
  map2(dysplastic_feature_labels, function(plot, label) {
    plot + theme_void() + ggtitle(label)
  })
p3f <- wrap_plots(p3f_plots, ncol = 2)
p3f
save_plot(p3f, file.path(figure_dir, "Figure3F_dysplastic_subtypes_diffusion.pdf"), 8, 7)
```

## Figure 3G: spatial dysplastic epithelial subtypes

```{r figure-3g-dysplastic-subtypes-spatial}
p3g_plots <- SpatialFeaturePlot(
  combined_final,
  features = dysplastic_features,
  images = main_spatial_example,
  combine = FALSE,
  keep.scale = "feature"
) |>
  map2(dysplastic_feature_labels, function(plot, label) {
    plot + ggtitle(label)
  })
p3g <- wrap_plots(p3g_plots, ncol = 2)
p3g
save_plot(p3g, file.path(figure_dir, "Figure3G_dysplastic_subtypes_spatial.pdf"), 8, 7)
```

## Figure 3H: ADT-CD8A

```{r figure-3h-adt-cd8a}
DefaultAssay(combined_final) <- "Spatial"
p3h_umap <- SCP::FeatureDimPlot(
  combined_final,
  reduction = "wnn.umap",
  features = "ADT-CD8A",
  show_stat = F,theme_use = SCP::theme_blank
) 
p3h_spatial <- SpatialFeaturePlot(
  combined_final,
  features = "ADT-CD8A",
  images = main_spatial_example,
  min.cutoff = 0
)
p3h <- p3h_umap / p3h_spatial
p3h
save_plot(p3h, file.path(figure_dir, "Figure3H_ADT_CD8A.pdf"), 5, 8)
DefaultAssay(combined_final) <- "Spatial"
```

## Figure 3I: ADT signals in paired pathology regions

```{r figure-3i-adt-region-comparison}
adt_region_tests <- readRDS(here("results", "tables", "adt_region_tests.rds"))
adt_region_data <- imap_dfr(adt_region_tests, function(result, feature) {
  result$wide |>
    pivot_longer(
      c(Normal, `Low-grade`),
      names_to = "Adenoma",
      values_to = "Signal"
    ) |>
    mutate(
      Feature = feature,
      Adenoma = recode(Adenoma, `Low-grade` = "Adenoma")
    )
}) |>
  mutate(
    Adenoma = factor(Adenoma, levels = c("Normal", "Adenoma")),
    Feature = recode(
      Feature,
      `ADT-CD8A` = "Normalized ADT-CD8A signal",
      `ADT-CD3E` = "Normalized ADT-CD3E signal"
    )
  )

p3i <- ggpubr::ggboxplot(
  adt_region_data,
  x = "Adenoma",
  y = "Signal",
  add = "point",
  fill = "Adenoma"
) +
  ggsci::scale_fill_aaas() +
  geom_line(aes(group = Sample_ID), alpha = 0.5) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    method.args = list(exact = FALSE),
    label = "p.format"
  ) +
  facet_wrap(~Feature, scales = "free_y") +
  labs(x = NULL, y = NULL) + NoLegend()
p3i
save_plot(p3i, file.path(figure_dir, "Figure3I_ADT_region_comparison.pdf"), 7, 4)
```

## Figure 3J: spatial IEX and SSIE co-occurrence scores

```{r figure-3j-iex-ssie-spatial}
DefaultAssay(combined_final) <- "Spatial"
p3j_plots <- SpatialFeaturePlot(
  combined_final,
  features = c("IEX", "SSIE"),
  images = "R4_large_2",
  combine = FALSE,
  keep.scale = "feature"
)
p3j <- wrap_plots(
  p3j_plots[[1]] + ggtitle("IEX score"),
  p3j_plots[[2]] + ggtitle("SSIE co-occurrence score"),
  ncol = 2
)
p3j
save_plot(p3j, file.path(figure_dir, "Figure3J_IEX_SSIE_spatial.pdf"), 6, 3)
```

## Figure 3K: SSIE co-occurrence score by cluster

```{r figure-3k-ssie-clusters}
p3k <- SCP::FeatureStatPlot(
  combined_final,
  stat.by = "SSIE",
  group.by = "combined_cluster",
  add_box = TRUE,
  assay = "Spatial4",
  palcolor = cluster_colors
)
p3k
save_plot(p3k, file.path(figure_dir, "Figure3K_SSIE_clusters.pdf"), 4, 2)
DefaultAssay(combined_final) <- "Spatial"
```

# Figure 4: GDF15, spatial association, and CD8 exclusion

## Figure 4A: GDF15 candidate intersection

```{r figure-4a-gdf15-overlap}
venn_data <- readRDS(here("results", "tables", "de_gdf15_venn.rds"))
p4a <- ggvenn::ggvenn(
  venn_data,
  show_stats = "c",
  set_name_size = 4,
  fill_color = c("#74C7D4", "#EE8C9B", "#F2C14E")
)
p4a
save_plot(p4a, file.path(figure_dir, "Figure4A_GDF15_overlap.pdf"), 5, 5)
```

## Figure 4B: GDF15 across clusters

```{r figure-4b-gdf15-clusters}
DefaultAssay(combined_final) <- "Spatial4"
p4b <- SCP::FeatureStatPlot(
  combined_final,
  stat.by = "GDF15",
  group.by = "combined_cluster",
  palcolor = cluster_colors, add_box = T
) +
  coord_flip() +
  theme_classic() +
  NoLegend()+
  labs(x = "Cluster", y = "Expression level")
p4b
save_plot(p4b, file.path(figure_dir, "Figure4B_GDF15_clusters.pdf"), 5, 5)
```

## Figure 4C: spatial GDF15

```{r figure-4c-spatial-gdf15}
p4c <- SpatialFeaturePlot(
  combined_final,
  features = "GDF15",
  images = advanced_examples,
  ncol = 1,
  keep.scale = "all"
)
p4c
save_plot(p4c, file.path(figure_dir, "Figure4C_spatial_GDF15.pdf"), 5, 8)
```

## Figure 4D: GDF15–senescence and GDF15–stemness compositions

```{r figure-4d-gdf15-signature-moran-composition}
gdf15_signature_columns <- c(
  `GDF15-Senescence` = "Cluster_Moran_GDF15_Senepy",
  `GDF15-Stemness` = "Cluster_Moran_GDF15_Stem_Signature"
)
p4d <- imap(gdf15_signature_columns, function(class_col, label) {
  size_plot <- plot_moran_composition(
    filter(combined_final@meta.data, Adenoma == "Low-grade"),
    class_col,
    "LargeSmall",
    c("Small", "Large"),
    posthoc = moran_composition_tests$largesmall[[class_col]]$posthoc
  ) + ggtitle(label)
  region_plot <- plot_moran_composition(
    combined_final@meta.data,
    class_col,
    "Adenoma",
    c("Normal", "Low-grade"),
    posthoc = moran_composition_tests$adenoma[[class_col]]$posthoc
  ) + ggtitle(label)
  size_plot | region_plot
}) |>
  wrap_plots(ncol = 1)
p4d
save_plot(p4d, file.path(figure_dir, "Figure4D_GDF15_signature_moran.pdf"), 8, 8)
```

## Figure 4E: representative GDF15-signature Moran maps

```{r figure-4e-gdf15-signature-moran-spatial}
combined_final_non_stroma = subset(combined_final,subset = Adenoma != "Stroma")
p4e_senescence <- SpatialDimPlot(
  combined_final_non_stroma,
  group.by = "Cluster_Moran_GDF15_Senepy",
  images = advanced_examples,
  cols = moran_colors,
  ncol = 1
) & NoLegend()
p4e_stemness <- SpatialDimPlot(
  combined_final_non_stroma,
  group.by = "Cluster_Moran_GDF15_Stem_Signature",
  images = advanced_examples,
  cols = moran_colors,
  ncol = 1
) & NoLegend()
p4e_pathology <- SpatialDimPlot(
  combined_final_non_stroma,
  group.by = "Adenoma",
  images = advanced_examples,
  cols = region_colors,
  ncol = 1
) & NoLegend()
p4e_clusters <- SpatialDimPlot(
  combined_final_non_stroma,
  group.by = "combined_cluster",
  images = advanced_examples,
  cols = cluster_colors,
  ncol = 1
) & NoLegend()
p4e <- p4e_senescence | p4e_stemness | p4e_pathology | p4e_clusters
p4e
save_plot(p4e, file.path(figure_dir, "Figure4E_GDF15_signature_moran_spatial.pdf"), 13, 7)
```

## Figure 4F: lag association before and after GDF15 adjustment

The inferential comparison is performed on Fisher-transformed correlations, so this panel plots `z_before` and `z_after` rather than raw rho.

```{r figure-4f-partial-spatial-correlation}
partial_spatial <- read_csv(
  here("results", "tables", "partial_spatial_correlation.csv"),
  show_col_types = FALSE
)
partial_spatial_test <- readRDS(
  here("results", "tables", "partial_spatial_correlation_test.rds")
)
partial_long <- partial_spatial |>
  dplyr::select(Sample_ID, group, z_before, z_after) |>
  pivot_longer(c(z_before, z_after), names_to = "Adjustment", values_to = "z") |>
  mutate(
    Adjustment = recode(Adjustment, z_before = "Before", z_after = "After"),
    Adjustment = factor(Adjustment, levels = c("Before", "After")),
    group = recode(group, Small = "Non-Adv", Large = "Advanced"),
    group = factor(group, levels = c("Non-Adv", "Advanced"))
  )
format_panel_p <- function(p) {
  if (is.na(p)) return("p = NA")
  if (p < 0.001) return("p < 0.001")
  paste0("p = ", formatC(p, format = "f", digits = 3))
}
partial_annotations <- partial_spatial_test$stratified |>
  transmute(
    group = recode(group, Small = "Non-Adv", Large = "Advanced"),
    group = factor(group, levels = c("Non-Adv", "Advanced")),
    label = map_chr(wilcox_p, format_panel_p)
  )
p4f <- sample_level_boxplot(partial_long, "Adjustment", "z") +
  facet_wrap(~group) +
  geom_text(
    data = partial_annotations,
    aes(x = 1.5, y = Inf, label = label),
    inherit.aes = FALSE,
    vjust = 1.5
  ) +
  labs(
    title = "Lag Senescence-Stemness correlation before/after GDF15 adjustment",
    subtitle = paste0(
      "Paired Wilcoxon ", format_panel_p(partial_spatial_test$paired_wilcox$p.value),
      " | Sign-flip permutation ", format_panel_p(partial_spatial_test$paired_perm_p)
    ),
    y = "Fisher z(rho)"
  ) + ggsci::scale_fill_aaas() + NoLegend()
p4f
save_plot(p4f, file.path(figure_dir, "Figure4F_partial_spatial_correlation.pdf"), 7, 4)
```

## Figure 4G: GDF15–CD8 Moran compositions

```{r figure-4g-gdf15-cd8-moran-composition}
gdf15_cd8_columns <- c(
  `GDF15-CD8A antibody` = "Cluster_Moran_GDF15_ADT_CD8A",
  `GDF15-CD8 T cell` = "Cluster_Moran_GDF15_CD8_T"
)
p4g <- imap(gdf15_cd8_columns, function(class_col, label) {
  size_plot <- plot_moran_composition(
    filter(combined_final@meta.data, Adenoma == "Low-grade"),
    class_col,
    "LargeSmall",
    c("Small", "Large"),
    posthoc = moran_composition_tests$largesmall[[class_col]]$posthoc
  ) + ggtitle(label)
  region_plot <- plot_moran_composition(
    combined_final@meta.data,
    class_col,
    "Adenoma",
    c("Normal", "Low-grade"),
    posthoc = moran_composition_tests$adenoma[[class_col]]$posthoc
  ) + ggtitle(label)
  size_plot | region_plot
}) |>
  wrap_plots(ncol = 1)
p4g
save_plot(p4g, file.path(figure_dir, "Figure4G_GDF15_CD8_moran.pdf"), 8, 8)
```

## Figure 4H: representative GDF15–CD8 Moran maps

```{r figure-4h-gdf15-cd8-moran-spatial}
p4h_adt <- SpatialDimPlot(
  combined_final,
  group.by = "Cluster_Moran_GDF15_ADT_CD8A",
  images = advanced_examples,
  cols = moran_colors,
  ncol = 1
) & NoLegend()
p4h_cell <- SpatialDimPlot(
  combined_final,
  group.by = "Cluster_Moran_GDF15_CD8_T",
  images = advanced_examples,
  cols = moran_colors,
  ncol = 1
) & NoLegend()
p4h <- p4h_adt | p4h_cell
p4h
save_plot(p4h, file.path(figure_dir, "Figure4H_GDF15_CD8_moran_spatial.pdf"), 8, 7)
```

## Figure 4I: CD8 density around GDF15-low and GDF15-high spots

```{r figure-4i-cd8-gdf15-high-low}
cd8_gdf15 <- read_csv(
  here("results", "tables", "cd8_density_gdf15_high_low.csv"),
  show_col_types = FALSE
)
cd8_gdf15_long <- cd8_gdf15 |>
  filter(!is.na(dens_High), !is.na(dens_Low)) |>
  dplyr::select(Sample_ID, LargeSmall, Adenoma, dens_Low, dens_High) |>
  pivot_longer(c(dens_Low, dens_High), names_to = "GDF15", values_to = "Density") |>
  mutate(
    GDF15 = recode(GDF15, dens_Low = "GDF15-Low", dens_High = "GDF15-High"),
    GDF15 = factor(GDF15, levels = c("GDF15-Low", "GDF15-High")),
    Adenoma = recode(Adenoma, `Low-grade` = "Adenoma"),
    Adenoma = factor(Adenoma, levels = c("Normal", "Adenoma"))
  )
p4i <- ggpubr::ggboxplot(
  cd8_gdf15_long,
  x = "GDF15",
  y = "Density",
  facet.by = "Adenoma",
  add = "point",
  width = 0.6,
  fill = "GDF15"
) +
  geom_line(aes(group = Sample_ID), alpha = 0.25) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    method.args = list(exact = TRUE),
    label = "p.format"
  ) +
  labs(x = NULL, y = "CD8 T cell density by antibody")+scale_fill_aaas() + NoLegend()
p4i
save_plot(p4i, file.path(figure_dir, "Figure4I_CD8_GDF15_high_low.pdf"), 6, 4)
```

## Figure 4J: CD8 density around GDF15-high spots by pathology

```{r figure-4j-cd8-region}
cd8_region <- read_csv(
  here("results", "tables", "cd8_density_region.csv"),
  show_col_types = FALSE
)
cd8_region_long <- cd8_region |>
  dplyr::select(Sample_ID, LargeSmall, dens_LGD, dens_Normal) |>
  pivot_longer(c(dens_LGD, dens_Normal), names_to = "Region", values_to = "Density") |>
  mutate(
    Region = recode(Region, dens_LGD = "Adenoma", dens_Normal = "Normal"),
    Region = factor(Region, levels = c("Normal", "Adenoma")),
    Advancement = recode(LargeSmall, Small = "Non-Adv", Large = "Advanced"),
    Advancement = factor(Advancement, levels = c("Non-Adv", "Advanced"))
  )
p4j <- ggpubr::ggboxplot(
  cd8_region_long,
  x = "Region",
  y = "Density",
  facet.by = "Advancement",
  add = "point",
  width = 0.6,
  fill = "Region"
) +
  geom_line(aes(group = Sample_ID), alpha = 0.25) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    method.args = list(exact = TRUE),
    label = "p.format"
  ) +
  labs(x = NULL, y = "CD8 T cell density by antibody around GDF15-high spots") +scale_fill_aaas() + NoLegend()
p4j
save_plot(p4j, file.path(figure_dir, "Figure4J_CD8_region.pdf"), 6, 4)
```
