---
title: "Supplementary figure generation"
output:
  html_document:
    toc: true
params:
  execute: false
---

This notebook contains the data-derived panels for Visium Supplementary Figures 1–7 in manuscript order.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(Seurat)
library(tidyverse)
library(qs)
library(here)
library(patchwork)
library(dittoSeq)
library(ggpubr)
source(here("R", "plotting.R"))
source(here("R", "spatial_statistics.R"))

combined_final <- qread(here("data", "derived", "visium_scored.qs"))
# for plotting with SCP, copy the Seurat5 data as a Seurat4 assay
# combined_final[["Spatial4"]] <- as(object = combined_final[["Spatial"]], Class = "Assay")
# DefaultAssay(combined_final) <- "Spatial4"
epithelial <- qread(here("data", "derived", "epithelial_diffusion_map.qs"))
figure_dir <- here("results", "figures", "supplementary")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

sample_order <- read_csv(here("config", "sample_manifest.csv"), show_col_types = FALSE) |>
  filter(include) |>
  pull(sample_id)

choose_reduction <- function(object, candidates = c("umap", "Xumap_", "XscANVI_")) {
  available <- candidates[candidates %in% names(object@reductions)]
  if (!length(available)) {
    stop("None of the requested reductions are present: ", paste(candidates, collapse = ", "))
  }
  available[[1]]
}
```

# Supplementary Figure 1: multimodal processing and annotation

## Supplementary Figure 1B: representative RNA and ADT spatial signals

Representative spatial maps show RNA expression and DSB-normalized ADT signals.

```{r sfigure-1b-rna-adt-spatial}
DefaultAssay(combined_final) <- "Spatial"
s1b_rna <- SpatialFeaturePlot(
  combined_final,
  features = c("CD8A", "CD3E", "PTPRC"),
  images = "R3_large_2",
  ncol = 3
)

DefaultAssay(combined_final) <- "ADT_dsb"
s1b_adt <- SpatialFeaturePlot(
  combined_final,
  features = c("ADT-CD8A", "ADT-CD3E", "ADT-PTPRC"),
  images = "R3_large_2",
  min.cutoff = 0,
  ncol = 3
)
s1b <- s1b_rna / s1b_adt
DefaultAssay(combined_final) <- "Spatial"
s1b
save_plot(s1b, file.path(figure_dir, "SFigure1B_RNA_ADT_spatial.pdf"), 9, 6)
```

## Supplementary Figure 1C: UMAP by sample

```{r sfigure-1c-umap-sample}
s1c <- SCP::CellDimPlot(combined_final,group.by = "Sample_ID", theme_use = SCP::theme_blank)
s1c
save_plot(s1c, file.path(figure_dir, "SFigure1C_UMAP_by_sample.pdf"), 7, 5)
```

## Supplementary Figure 1D–E: pathology and cluster compositions

```{r sfigure-1de-compositions}
s1d <- dittoSeq::dittoBarPlot(
  combined_final,
  var = "Adenoma",
  group.by = "combined_cluster",
  color.panel = unname(region_colors),
  retain.factor.levels = TRUE,
  ylab = "Relative abundance"
)
s1e <- dittoSeq::dittoBarPlot(
  combined_final,
  var = "combined_cluster",
  group.by = "Adenoma",
  split.by = "LargeSmall",
  color.panel = unname(cluster_colors),
  retain.factor.levels = TRUE,
  ylab = "Relative abundance"
)
s1d|s1e
save_plot(s1d, file.path(figure_dir, "SFigure1D_pathology_by_cluster.pdf"), 5, 4)
save_plot(s1e, file.path(figure_dir, "SFigure1E_clusters_by_pathology.pdf"), 7, 4)
```

# Supplementary Figure 2: all Visium spatial maps

```{r sfigure-2-all-sections,fig.width=16,fig.height=10}
s2a <- plot_spatial_samples(
  combined_final,
  group_by = "combined_cluster",
  samples = sample_order,
  colors = cluster_colors,
  ncol = 6,
  point_size = 1.3
)
s2b <- plot_spatial_samples(
  combined_final,
  group_by = "Adenoma",
  samples = sample_order,
  colors = region_colors,
  ncol = 6,
  point_size = 1.3
)
s2a/s2b & NoLegend()
save_plot(s2a, file.path(figure_dir, "SFigure2A_all_sections_clusters.pdf"), 16, 10)
save_plot(s2b, file.path(figure_dir, "SFigure2B_all_sections_pathology.pdf"), 16, 10)
```

# Supplementary Figure 3: spatial-cluster characterization

## Supplementary Figure 3A: RNA and ADT markers

```{r sfigure-3a-rna-adt-markers,fig.width=12,fig.height=7}
picked_marker_gene <- c(
  "WFDC2", "CLCA1", "TFF3",
  "IGKC", "COL3A1",
  "ASCL2", "HMGB2", "LGR5", "MKI67",
  "SLC26A3", "CDKN2B", "CEACAM7",
  "CEACAM6", "CDKN1A", "AXIN2",
  "DES", "GREM1",
  "CA2", "ADH1C", "CHGA",
  "TRAC", "MS4A1", "CCL19",
  "REG1A", "XBP1", "OLFM4"
)
adt_markers <- c(
  "ADT-EPCAM", "ADT-KRT5", "ADT-PECAM1", "ADT-VIM", "ADT-ACTA2",
  "ADT-PTPRC", "ADT-PTPRC.2", "ADT-CD3E", "ADT-CD4", "ADT-CD8A",
  "ADT-CD19", "ADT-MS4A1", "ADT-CD68", "ADT-CD14", "ADT-CD163",
  "ADT-ITGAX", "ADT-CD40", "ADT-CCR7", "ADT-CXCR5"
)

DefaultAssay(combined_final) <- "Spatial"
s3a_rna <- DotPlot(
  combined_final,
  features = picked_marker_gene,
  group.by = "combined_cluster",
  col.min = -2,
  col.max = 2,
  scale = TRUE,
  cols = c("grey90", "red")
) +
  scale_colour_gradient2(low = "#441153", mid = "grey90", high = "#3CB67B") +
  coord_flip() +
  theme_classic() +
  theme(axis.title = element_blank()) +
  ggtitle("RNA")

DefaultAssay(combined_final) <- "ADT_dsb_by_slide"
s3a_adt <- DotPlot(
  combined_final,
  features = adt_markers,
  group.by = "combined_cluster",
  col.min = -2,
  col.max = 2,
  scale = TRUE,
  cols = c("grey90", "red")
) +
  scale_colour_gradient2(low = "#441153", mid = "grey90", high = "#3CB67B") +
  scale_x_discrete(labels = function(x) recode(
    x,
    "ADT-PTPRC" = "ADT-CD45RA",
    "ADT-PTPRC.2" = "ADT-CD45RO",
    .default = x
  )) +
  coord_flip() +
  theme_classic() +
  theme(axis.title = element_blank()) +
  ggtitle("ADT")

s3a <- s3a_rna | s3a_adt
DefaultAssay(combined_final) <- "Spatial"
s3a
save_plot(s3a, file.path(figure_dir, "SFigure3A_RNA_ADT_markers.pdf"), 12, 7)

```

## Supplementary Figure 3C: integrated gene-set enrichment

```{r sfigure-3c-irgsea-heatmap,fig.height=8}
library(irGSEA)

result.dge = readRDS(here("data", "derived", "result.dge.rds"))

significant_sets <- result.dge[["RRA"]] |>
  dplyr::filter(pvalue < 0.05, !grepl("WORSCHECH", Name)) |>
  distinct(Name) |>
  pull(Name)
s3c <- irGSEA::irGSEA.heatmap(
  object = result.dge,
  method = "RRA",
  top = 57,
  cluster.color = unname(cluster_colors),
  show.geneset = significant_sets
)
s3c
save_plot(s3c, file.path(figure_dir, "SFigure3C_irGSEA_heatmap.pdf"), 9, 6)
```

## Supplementary Figure 3D: signature activity by cluster

```{r sfigure-3d-signature-violins}
signature_features <- c(
  "Senepy", "Stem_Signature", "E2F_TARGETS", "MYC_TARGETS",
  "G2M_CHECKPOINT", "TGF_BETA_SIGNALING", "WNT_BETA_CATENIN_SIGNALING",
  "P53_PATHWAY", "INFLAMMATORY_RESPONSE", "IEX"
)
s3d <- SCP::FeatureStatPlot(
  combined_final,
  stat.by = signature_features,
  group.by = "combined_cluster",
  palcolor  = cluster_colors,
  add_box = T,
  stack = TRUE,
  assay = "Spatial4"
)
s3d
save_plot(s3d, file.path(figure_dir, "SFigure3D_signature_activity.pdf"), 9, 10)
```

## Supplementary Figure 3E–F: cell-cycle classifications

```{r sfigure-3ef-cell-cycle}
s3e <- dittoSeq::dittoBarPlot(
  object = combined_final,
  var = "cell_cycle_seurat",
  group.by = "combined_cluster",
  retain.factor.levels = TRUE,
  ylab = "Relative abundance"
)
s3f <- dittoSeq::dittoBarPlot(
  object = combined_final,
  var = "ccAFv2",
  group.by = "combined_cluster",
  retain.factor.levels = TRUE,
  ylab = "Relative abundance"
)
s3e|s3f
save_plot(s3e, file.path(figure_dir, "SFigure3E_cell_cycle_Seurat.pdf"), 5, 3)
save_plot(s3f, file.path(figure_dir, "SFigure3F_cell_cycle_ccAFv2.pdf"), 5, 3)
```

## Supplementary Figure 3G: sample-level SenePy comparisons

```{r sfigure-3g-senepy-cluster-comparisons}
continuous_tests <- readRDS(here("results", "tables", "continuous_score_tests.rds"))

paired_score_plot <- function(test_result, levels, colors) {
  plot_data <- test_result$wide |>
    pivot_longer(all_of(levels), names_to = "Cluster", values_to = "Senepy") |>
    mutate(Cluster = factor(paste0("C", Cluster), levels = paste0("C", levels)))

  ggpubr::ggboxplot(
    plot_data,
    x = "Cluster",
    y = "Senepy",
    add = "point",
    width = 0.6,
    fill = "Cluster"
  ) +
    scale_fill_manual(values = colors) +
    geom_line(aes(group = Sample_ID), alpha = 0.25) +
    ggpubr::stat_compare_means(
      method = "wilcox.test",
      paired = TRUE,
      method.args = list(exact = FALSE),
      label = "p.format"
    ) +
    labs(x = NULL, y = "Sample-level median SenePy score")
}

s3g_c2c4 <- paired_score_plot(
  continuous_tests$senepy_c4_vs_c2,
  c("2", "4"),
  unname(cluster_colors[c("2", "4")])
)
s3g_c3c4 <- paired_score_plot(
  continuous_tests$senepy_c4_vs_c3,
  c("3", "4"),
  unname(cluster_colors[c("3", "4")])
)
s3g <- (s3g_c2c4 | s3g_c3c4) & coord_cartesian(ylim = c(0.03, 0.11))
s3g # shows unadjusted p-value
save_plot(s3g, file.path(figure_dir, "SFigure3G_SenePy_cluster_comparisons.pdf"), 6, 4)
```

# Supplementary Figure 4: reference-guided spatial mapping

```{r sfigure-4-load-reference}
marteau_reference <- qread(
  here("data", "reference", "marteau_TA_updated.qs")
)

reference_reduction <- choose_reduction(marteau_reference)
dysplastic_cells <- rownames(marteau_reference@meta.data)[
  grepl("^Dysplastic", marteau_reference$my_annotation)
]
dysplastic_reference <- subset(marteau_reference, cells = dysplastic_cells)
dysplastic_reference = RunUMAP(dysplastic_reference, reduction = "XscANVI_",dims = 1:10)
dysplastic_reference = subset(dysplastic_reference,subset = my_annotationV2 != "Dysplastic_Normal")
dysplastic_reference$my_annotationV2 = recode(dysplastic_reference$my_annotationV2,
                                              "Dysplastic_Absorptive_like" = "STYL2+ Dysplastic cell",
                                              "Dysplastic_Paneth_like" = "DEFA5+ Dysplastic cell",
                                              "Dysplastic_Proliferating" = "MKI67+ Dysplastic cell",
                                              "Dysplastic_Senescent_like" = "LGALS3+ Dysplastic cell",
                                              "Dysplastic_Stem_like" = "LGR5+ Dysplastic cell",
                                              "Dysplastic_Tuft_like" = "AVIL+ Dysplastic cell")

table(dysplastic_reference$my_annotationV2)
dysplastic_reference$my_annotationV2 = factor(dysplastic_reference$my_annotationV2,levels = c("LGR5+ Dysplastic cell","STYL2+ Dysplastic cell","LGALS3+ Dysplastic cell",
                                                              "MKI67+ Dysplastic cell","AVIL+ Dysplastic cell","DEFA5+ Dysplastic cell"))
```

## Supplementary Figure 4A: representative Visium maps

```{r sfigure-4a-spatial-summary,fig.width=11,fig.height=7}
s4a_sample <- "R4_large_1"
s4a_pathology <- SpatialDimPlot(
  combined_final,
  group.by = "Adenoma",
  images = s4a_sample,
  cols = region_colors
)
s4a_he <- SpatialDimPlot(
  combined_final,
  group.by = "Adenoma",
  images = s4a_sample,
  cols = region_colors,
  alpha = 0
) + NoLegend()
s4a_scores <- SpatialFeaturePlot(
  combined_final,
  features = c("Senepy", "Stem_Signature"),
  images = s4a_sample,
  ncol = 1
)
s4a_moran <- SpatialDimPlot(
  combined_final,
  group.by = "Cluster_Moran_Senepy_Stem_Signature",
  images = s4a_sample,
  cols = moran_colors
)
s4a_clusters <- SpatialDimPlot(
  combined_final,
  group.by = "combined_cluster",
  images = s4a_sample,
  cols = cluster_colors
)
s4a <- wrap_plots(
  s4a_pathology, s4a_scores[[1]], s4a_moran,
  s4a_he, s4a_scores[[2]], s4a_clusters,
  ncol = 3
)
s4a
save_plot(s4a, file.path(figure_dir, "SFigure4A_spatial_summary.pdf"), 11, 7)
```

## Supplementary Figure 4B: CytoTRACE2

```{r sfigure-4b-cytotrace}
cytotrace_cells <- rownames(combined_final@meta.data)[!is.na(combined_final$CytoTRACE2_Score)]
cytotrace_epithelium <- subset(combined_final, cells = cytotrace_cells)
s4b_umap <- SCP::FeatureDimPlot(
  cytotrace_epithelium,
  reduction = "wnn.umap",
  features = "CytoTRACE2_Score",
  show_stat = F,theme_use = SCP::theme_blank
) 
s4b_violin <- SCP::FeatureStatPlot(
  cytotrace_epithelium,
  stat.by = "CytoTRACE2_Score",
  group.by = "combined_cluster",
  palcolor = cluster_colors,
  add_box = T,assay = "Spatial4"
  
) + coord_flip()
s4b <- s4b_umap | s4b_violin
s4b
save_plot(s4b, file.path(figure_dir, "SFigure4B_CytoTRACE2.pdf"), 8, 4)
```

## Supplementary Figure 4C–H: Cell2location reference

```{r sfigure-4c-reference-cell-types}
s4c <- SCP::CellDimPlot(
  marteau_reference,
  reduction = "umap",
  group.by = "cell_type_middle",
  theme_use = SCP::theme_blank
)
s4c
save_plot(s4c, file.path(figure_dir, "SFigure4C_reference_cell_types.pdf"), 7, 6)
```

```{r sfigure-4d-reference-scores,fig.width=10,fig.height=6}
s4d <- SCP::FeatureDimPlot(
  marteau_reference,
  reduction = reference_reduction,
  features = c("Stem_Signature_UCell", "Universal_score"),
  ncol = 2,
  keep_scale = "feature",show_stat = F,theme_use = SCP::theme_blank,
  raster = T,raster.dpi = c(300,300)
) 
s4d
save_plot(s4d, file.path(figure_dir, "SFigure4D_reference_scores.pdf"), 8, 4)
```

```{r sfigure-4e-dysplastic-annotations}
s4e <- SCP::CellDimPlot(
  dysplastic_reference,
  reduction = reference_reduction,
  group.by = "my_annotationV2",
  theme_use = SCP::theme_blank,
  show_stat = F
) 
s4e
save_plot(s4e, file.path(figure_dir, "SFigure4E_dysplastic_annotations.pdf"), 7, 5)
```

```{r sfigure-4f-dysplastic-scores}
s4f <- SCP::FeatureDimPlot(
  dysplastic_reference,
  reduction = reference_reduction,
  features = c("epithelial_cell_score_0", "Stem_Signature_UCell"),
  ncol = 1,
  keep_scale = "feature",
  show_stat = F,
  theme_use = SCP::theme_blank
)
s4f
save_plot(s4f, file.path(figure_dir, "SFigure4F_dysplastic_scores.pdf"), 5, 8)
```

```{r sfigure-4g-dysplastic-markers}
reference_markers <- c(
  "LGR5", "SMOC2", "LEFTY1", "SYTL2", "ABCC3", "PPARG",
  "FABP1", "LGALS3", "MMP7", "MKI67", "TOP2A", "UBE2C",
  "AVIL", "POU2F3", "HPGDS", "DEFA5", "DEFA6", "REG1B"
)

s4g = DotPlot(
  dysplastic_reference,
  features = reference_markers,
  group.by = "my_annotationV2",
  col.min = -2,
  col.max = 2,
  scale = TRUE,
  cols = c("grey90", "red")
) +
  scale_colour_gradient2(low = "#441153", mid = "grey90", high = "#3CB67B") +
  coord_flip() +
  theme_classic() +
  theme(axis.title = element_blank()) 
s4g

save_plot(s4g, file.path(figure_dir, "SFigure4G_dysplastic_markers.pdf"), 8, 5)
```

```{r sfigure-4h-dysplastic-violins,fig.width=8,fig.height=7}
s4h_features <- c(
  "LGR5", "OLFM4", "CDKN1A", "Stem_Signature_UCell", "epithelial_cell_score_0"
)
s4h <- SCP::FeatureStatPlot(
  dysplastic_reference,
  stat.by = s4h_features,
  group.by = "my_annotationV2",
  add_box = T,
  stack = TRUE
)
s4h
save_plot(s4h, file.path(figure_dir, "SFigure4H_dysplastic_violins.pdf"), 8, 7)
```

## Supplementary Figure 4I: spatial Cell2location abundance and ADT-CD8A

```{r sfigure-4i-spatial-deconvolution,fig.width=13,fig.height=7}
s4i_subtypes <- SpatialFeaturePlot(
  combined_final,
  features = c(
    "Dysplastic_Senescent_like", "Dysplastic_Stem_like",
    "Dysplastic_Tuft_like", "Dysplastic_Proliferating"
  ),
  images = "R4_large_1",
  combine = FALSE,
  keep.scale = "feature"
) |>
  map2(
    c(
      "LGALS3+ Dysplastic cell", "LGR5+ Dysplastic cell",
      "AVIL+ Dysplastic cell", "MKI67+ Dysplastic cell"
    ),
    function(plot, label) plot + ggtitle(label)
  )
s4i_three_way <- three_way_spatial_plot(
  combined_final,
  features = c("Dysplastic_Epi", "Normal_Epi", "CD8_T_cells"),
  images = "R4_large_1",
  colors = c("#E41A1C", "#377EB8", "#5ce828"),
  legend_titles = c("Dysplastic Epi", "Normal Epi", "CD8 T cells")
)
DefaultAssay(combined_final) <- "ADT_dsb_by_slide"
s4i_adt <- SpatialFeaturePlot(
  combined_final,
  features = "ADT-CD8A",
  images = "R4_large_1",
  min.cutoff = 0
)
s4i <- wrap_plots(
  s4i_subtypes[[1]], s4i_subtypes[[2]], s4i_subtypes[[3]], s4i_subtypes[[4]],
  s4i_three_way, s4i_adt,
  ncol = 3
)
s4i
save_plot(s4i, file.path(figure_dir, "SFigure4I_spatial_deconvolution.pdf"), 13, 7)
DefaultAssay(combined_final) <- "Spatial"
```

## Supplementary Figure 4J: ADT signals in paired pathology regions

```{r sfigure-4j-adt-region-comparison}
adt_region_tests <- readRDS(here("results", "tables", "adt_region_tests.rds"))
adt_region_data <- imap_dfr(adt_region_tests, function(result, feature) {
  result$wide |>
    pivot_longer(
      c(Normal, `Low-grade`),
      names_to = "Adenoma",
      values_to = "Signal"
    ) |>
    mutate(
      Feature = feature,
      Adenoma = recode(Adenoma, `Low-grade` = "Adenoma")
    )
}) |>
  mutate(
    Adenoma = factor(Adenoma, levels = c("Normal", "Adenoma")),
    Feature = recode(
      Feature,
      `ADT-CD8A` = "Normalized ADT-CD8A signal",
      `ADT-CD3E` = "Normalized ADT-CD3E signal"
    )
  )

s4j <- ggpubr::ggboxplot(
  adt_region_data,
  x = "Adenoma",
  y = "Signal",
  add = "point",
  fill = "Adenoma"
) +
  ggsci::scale_fill_aaas() +
  geom_line(aes(group = Sample_ID), alpha = 0.5) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    method.args = list(exact = FALSE),
    label = "p.format"
  ) +
  facet_wrap(~Feature, scales = "free_y") +
  labs(x = NULL, y = NULL) + NoLegend()
s4j
save_plot(s4j, file.path(figure_dir, "SFigure4J_ADT_region_comparison.pdf"), 7, 4)
```

# Supplementary Figure 5: Cell2location-inferred cell types

```{r sfigure-5-cell2location-abundance,fig.width=14,fig.height=20}
cell2location_features <- c(
  "Normal.Colonocyte", "Normal.Colonocyte.BEST4", "Normal.Crypt.cell", "Normal.Goblet",
  "Normal.TA", "Tuft", "Enteroendocrine", "Dysplastic_Absorptive_like",
  "Dysplastic_Paneth_like", "Dysplastic_Proliferating", "Dysplastic_Senescent_like",
  "Dysplastic_Stem_like", "Dysplastic_Tuft_like", "CD4.T", "Treg", "CD8.T",
  "gamma.delta", "NK", "B.cell.activated", "GC.B.cell", "Plasma.IgG", "Plasma.IgA",
  "Fibroblast", "Endothelial", "Macrophage", "Macrophage_Monocyte", "Mast.cell", "cDC"
)
missing_cell2location <- setdiff(cell2location_features, colnames(combined_final@meta.data))
if (length(missing_cell2location)) {
  stop("Missing Cell2location columns: ", paste(missing_cell2location, collapse = ", "))
}
s5 <- SCP::FeatureDimPlot(
  combined_final,
  reduction = "wnn.umap",
  features = cell2location_features,
  ncol = 4,
  keep_scale = "feature",
  show_stat = F,pt.size = 2,
  theme_use = SCP::theme_blank,raster = T,raster.dpi = c(600,600)
)
s5
save_plot(s5, file.path(figure_dir, "SFigure5_cell2location_abundance.pdf"), 14, 20)
```

# Supplementary Figure 6: independent scRNA-seq validation

The two epithelial validation cohorts are distinct from the Marteau Cell2location reference.

```{r sfigure-6-load-validation}
validation_objects <- qread(
  here("data", "derived", "single_cell_validation_full.qs")
)
validation_epithelium <- qread(
  here("data", "derived", "single_cell_validation.qs")
)
validation_non_epi <- qread((here("data", "derived", "single_cell_validation_non_epi.qs")))
```

```{r sfigure-6-validation-panels,fig.width=14,fig.height=14}
s6_panels <- imap(validation_objects, function(object, cohort) {
  epithelial_object <- validation_epithelium[[cohort]]
  full_reduction <- choose_reduction(object)
  epithelial_reduction <- choose_reduction(epithelial_object)

  annotation_plot <- SCP::CellDimPlot(
    object,
    reduction = full_reduction,
    group.by = "Cell_Type",
    show_stat = F,theme_use = SCP::theme_blank
  )  + ggtitle(paste(cohort, "cell identities"))
  senescence_plot <- SCP::FeatureDimPlot(
    object,
    reduction = full_reduction,
    features = "epithelial_cell_score_0",
    show_stat = F,theme_use = SCP::theme_blank
  ) 
  stemness_plot <- SCP::FeatureDimPlot(
    object,
    reduction = full_reduction,
    features = "Stem_Signature",
    show_stat = F,theme_use = SCP::theme_blank
  )
  subclusters <- SCP::CellDimPlot(
    epithelial_object,
    reduction = epithelial_reduction,
    group.by = "seurat_clusters",
    show_stat = F,theme_use = SCP::theme_blank
  ) 
  violins <- SCP::FeatureStatPlot(
    epithelial_object,
    stat.by = c(
      "epithelial_cell_score_0", "WNT_BETA_CATENIN_SIGNALING", "Stem_Signature", "IEX",
      "CDKN1A", "TP53", "GLB1", "LGR5", "OLFM4"
    ),
    group.by = "seurat_clusters",
    stack = TRUE,
    add_box = T
  )
  (annotation_plot | senescence_plot | stemness_plot) /
    (subclusters | violins)
}) |>
  wrap_plots(ncol = 1)
s6_panels
save_plot(s6_panels, file.path(figure_dir, "SFigure6_single_cell_validation.pdf"), 14, 14)
```


# Supplementary Figure 6k: Senepy comparison between adenoma cells and non-epi cells

```{r sfigure-6k}
ASC_senepy = validation_epithelium$discovery@meta.data %>% dplyr::select(c("Cell_Type","Universal_score"))
ASC_senepy = rbind(ASC_senepy, validation_epithelium$validation@meta.data %>% dplyr::select(c("Cell_Type","Universal_score")))
Non_senepy = validation_non_epi@meta.data %>% dplyr::select(c("Cell_Type","Universal_score"))

senepy_compare = rbind(ASC_senepy,Non_senepy)

s6k = ggviolin(senepy_compare,x = "Cell_Type",y = "Universal_score",add = c("boxplot"),fill = "Cell_Type",palette = as.character(cluster_colors)) + stat_compare_means(comparisons = list(
  c("ASC","T"),
  c("ASC","PLA"),
  c("ASC","MAS"),
  c("ASC","MYE"),
  c("ASC","FIB"),
  c("ASC","B"),
  c("ASC","END")
  ), symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("****", "***", "**", "*", "ns"))
) + NoLegend()
s6k
save_plot(s6k, file.path(figure_dir, "SFigure6K_Senepy_comparison.pdf"), 8, 4)
```


# Supplementary Figure 7: GDF15 validation and CD8 neighborhoods

## Supplementary Figure 7A: sample-level mean GDF15

Each point is one sample-level mean. The advancement comparison is unpaired; the stemness and senescence-region comparisons are paired within sample.

```{r sfigure-7a-gdf15-advancement}
DefaultAssay(combined_final) = "Spatial"
combined_final_epi <- subset(combined_final, idents = c("2", "4"))
gdf15_adv_sample <- FetchData(
  combined_final_epi,
  vars = c("GDF15", "Sample_ID", "LargeSmall", "combined_cluster")
) |>
  dplyr::filter(combined_cluster %in% c("2", "4"), !is.na(LargeSmall)) |>
  group_by(Sample_ID, LargeSmall) |>
  summarise(
    mean_GDF15 = mean(GDF15, na.rm = TRUE),
    n_spots = n(),
    .groups = "drop"
  ) |>
  mutate(LargeSmall = factor(LargeSmall, levels = c("Small", "Large")))

gdf15_adv_test <- wilcox.test(
  mean_GDF15 ~ LargeSmall,
  data = gdf15_adv_sample,
  alternative = "two.sided",
  exact = FALSE
)

gdf15_sample <- ggpubr::ggboxplot(
  gdf15_adv_sample,
  x = "LargeSmall",
  y = "mean_GDF15",
  add = "point",
  width = 0.6,
  fill = "LargeSmall"
) +
  scale_fill_manual(values = size_colors) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = FALSE,
    method.args = list(exact = FALSE),
    label = "p.format"
  ) +
  labs(x = NULL, y = "Sample-level mean GDF15 expression")
```

```{r sfigure-7a-gdf15-stemness}
stem_cutoffs <- quantile(
  combined_final_epi$Stem_Signature,
  probs = c(0.25, 0.75),
  na.rm = TRUE
)
combined_final_epi$StemGroup <- case_when(
  combined_final_epi$Stem_Signature <= stem_cutoffs[[1]] ~ "Low",
  combined_final_epi$Stem_Signature >= stem_cutoffs[[2]] ~ "High",
  TRUE ~ "Mid"
)
gdf15_stem_sample <- FetchData(
  combined_final_epi,
  vars = c("GDF15", "Sample_ID", "StemGroup")
) |>
  filter(StemGroup %in% c("High", "Low")) |>
  group_by(Sample_ID, StemGroup) |>
  summarise(mean_GDF15 = mean(GDF15, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = StemGroup, values_from = mean_GDF15) |>
  filter(!is.na(High), !is.na(Low))

gdf15_stemtest <- wilcox.test(
  gdf15_stem_sample$High,
  gdf15_stem_sample$Low,
  paired = TRUE,
  alternative = "two.sided"
)

gdf15_stem_long <- gdf15_stem_sample |>
  pivot_longer(c(Low, High), names_to = "Stem_group", values_to = "GDF15") |>
  mutate(Stem_group = factor(Stem_group, levels = c("Low", "High")))
gdf15_stem <- ggpubr::ggboxplot(
  gdf15_stem_long,
  x = "Stem_group",
  y = "GDF15",
  add = "point",
  width = 0.6,
  fill = "Stem_group"
) +
  scale_fill_manual(values = c(Low = "#0652DD", High = "#ff5252")) +
  geom_line(aes(group = Sample_ID), alpha = 0.25) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format"
  ) +
  labs(x = NULL, y = "Sample-level mean GDF15 expression")
```

```{r sfigure-7a-gdf15-senescence}
nonstromal <- subset(combined_final, subset = Adenoma != "Stroma")
senescence_cutoff <- mean(nonstromal$Senepy, na.rm = TRUE) +
  2 * sd(nonstromal$Senepy, na.rm = TRUE)
nonstromal$Senepy_pos <- if_else(nonstromal$Senepy > senescence_cutoff, "Yes", "No")

gdf15_sene_sample <- FetchData(
  nonstromal,
  vars = c("GDF15", "Sample_ID", "Adenoma", "Senepy_pos")
) |>
  filter(Senepy_pos == "Yes", Adenoma %in% c("Low-grade", "Normal")) |>
  group_by(Sample_ID, Adenoma) |>
  summarise(mean_GDF15 = mean(GDF15, na.rm = TRUE), n_spots = n(), .groups = "drop") |>
  pivot_wider(id_cols = Sample_ID,names_from = Adenoma,values_from = mean_GDF15) |>
  filter(!is.na(`Low-grade`), !is.na(Normal))

gdf15_sene_test <- wilcox.test(
  gdf15_sene_sample$`Low-grade`,
  gdf15_sene_sample$Normal,
  paired = TRUE,
  alternative = "two.sided",
  exact = FALSE
)

gdf15_sene_long <- gdf15_sene_sample |>
  pivot_longer(c(Normal, `Low-grade`), names_to = "Region", values_to = "GDF15") |>
  mutate(
    Region = recode(Region, `Low-grade` = "Adenoma"),
    Region = factor(Region, levels = c("Normal", "Adenoma"))
  )
gdf15_sen <- ggpubr::ggboxplot(
  gdf15_sene_long,
  x = "Region",
  y = "GDF15",
  add = "point",
  width = 0.6,
  fill = "Region"
) +
  scale_fill_manual(values = c(Normal = "#0652DD", Adenoma = "#ff5252")) +
  geom_line(aes(group = Sample_ID), alpha = 0.25) +
  ggpubr::stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    method.args = list(exact = FALSE),
    label = "p.format"
  ) +
  labs(x = NULL, y = "Sample-level mean GDF15 expression")
```

```{r sfigure-7a-summary}
gdf15_adv_test
gdf15_stemtest
gdf15_sene_test
s7a <- gdf15_sample | gdf15_stem | gdf15_sen
s7a
save_plot(s7a, file.path(figure_dir, "SFigure7A_GDF15_sample_level_means.pdf"), 8, 4)
```

## Supplementary Figure 7B: GDF15 across Visium spots

```{r sfigure-7b-visium-gdf15}
DefaultAssay(combined_final) <- "Spatial"
s7b <- SCP::FeatureDimPlot(
  combined_final,
  reduction = "wnn.umap",
  features = "GDF15",
  theme_use = SCP::theme_blank,show_stat = F,assay = "Spatial4"
) 
s7b
save_plot(s7b, file.path(figure_dir, "SFigure7B_Visium_GDF15.pdf"), 5, 4)
```

## Supplementary Figure 7C–F: GDF15 in public scRNA-seq data

```{r sfigure-7cdef-single-cell-gdf15,fig.width=12,fig.height=8}
s7cdf <- imap(validation_objects, function(object, cohort) {
  epithelial_object <- validation_epithelium[[cohort]]
  full_reduction <- choose_reduction(object)
  epithelial_reduction <- choose_reduction(epithelial_object)

  annotation <- SCP::CellDimPlot(
    object,
    reduction = full_reduction,
    group.by = "Cell_Type",theme_use = SCP::theme_blank,show_stat = F
  ) 
  gdf15_full <- SCP::FeatureDimPlot(
    object,
    reduction = full_reduction,
    features = "GDF15",theme_use = SCP::theme_blank,show_stat = F
  ) 
  gdf15_subclusters <- SCP::FeatureStatPlot(
    epithelial_object,
    stat.by = "GDF15",
    group.by = "seurat_clusters",
    add_box = T
  )
  annotation | gdf15_full | gdf15_subclusters
}) |>
  wrap_plots(ncol = 1)
s7cdf
save_plot(s7cdf, file.path(figure_dir, "SFigure7C_F_single_cell_GDF15.pdf"), 12, 8)
```
## Supplementary Figure 7E: Non epi GDF15

```{r sfigure-7e-non-epi-GDF15}
annotation <- SCP::CellDimPlot(
    validation_non_epi,
    group.by = "Cell_Type",theme_use = SCP::theme_blank,show_stat = F
  ) 
gdf15_full <- SCP::FeatureDimPlot(
    validation_non_epi,
    features = "GDF15",theme_use = SCP::theme_blank,show_stat = F
  ) 
s7e = annotation+gdf15_full
s7e
save_plot(s7e, file.path(figure_dir, "SFigure7E_single_cell_GDF15.pdf"), 6, 8)
```



## Supplementary Figure 7G: representative local Moran maps

```{r sfigure-7g-local-moran-spatial,fig.width=15,fig.height=4}
s7g_sample <- "R4_small_2"
s7g <- wrap_plots(
  SpatialDimPlot(
    combined_final,
    group.by = "Cluster_Moran_GDF15_Senepy",
    images = s7g_sample,
    cols = moran_colors
  ),
  SpatialDimPlot(
    combined_final,
    group.by = "Cluster_Moran_GDF15_Stem_Signature",
    images = s7g_sample,
    cols = moran_colors
  ),
  SpatialDimPlot(
    combined_final,
    group.by = "Adenoma",
    images = s7g_sample,
    cols = region_colors
  ),
  SpatialDimPlot(
    combined_final,
    group.by = "combined_cluster",
    images = s7g_sample,
    cols = cluster_colors
  ),
  SpatialDimPlot(
    combined_final,
    group.by = "Cluster_Moran_GDF15_ADT_CD8A",
    images = s7g_sample,
    cols = moran_colors
  ),
  SpatialDimPlot(
    combined_final,
    group.by = "Cluster_Moran_GDF15_CD8_T",
    images = s7g_sample,
    cols = moran_colors
  ),
  ncol = 6
) & NoLegend()
s7g
save_plot(s7g, file.path(figure_dir, "SFigure7G_local_moran_spatial.pdf"), 15, 4)
```

## Supplementary Figure 7H: radius-based CD8 neighborhood illustration

The illustration uses the two-spot-unit neighborhood radius applied in the analysis.

```{r sfigure-7h-cd8-neighborhood-illustration}
cd8_demo_data <- prepare_cd8_neighborhood_data(
  combined_final,
  adt_assay = "ADT_dsb_by_slide"
) |>
  filter(Sample_ID == "R4_large_1") |>
  mutate(
    GDF15_group = factor(GDF15_group, levels = c("Low", "High")),
    Region = recode(Adenoma, `Low-grade` = "Adenoma"),
    spot_id = barcode
  )

spot_spacing <- median(
  RANN::nn2(
    as.matrix(select(cd8_demo_data, array_col, array_row)),
    as.matrix(select(cd8_demo_data, array_col, array_row)),
    k = 2
  )$nn.dists[, 2],
  na.rm = TRUE
)
radius <- 2 * spot_spacing

pick_example_spot <- function(
    data,
    region,
    gdf15_group,
    prefer_cd8 = TRUE
) {
  candidates <- data |>
    filter(Region == region, as.character(GDF15_group) == gdf15_group)
  if (!nrow(candidates)) {
    stop("No candidate center for ", region, " / GDF15-", gdf15_group)
  }

  center_x <- median(candidates$array_col, na.rm = TRUE)
  center_y <- median(candidates$array_row, na.rm = TRUE)
  if (!prefer_cd8) {
    return(
      candidates |>
        mutate(center_distance = (array_col - center_x)^2 + (array_row - center_y)^2) |>
        slice_min(center_distance, n = 1, with_ties = FALSE)
    )
  }

  density_at_center <- function(index) {
    candidate <- candidates[index, , drop = FALSE]
    component <- data |>
      filter(tissue_island == candidate$tissue_island)
    inside <- (component$array_col - candidate$array_col)^2 +
      (component$array_row - candidate$array_row)^2 <= radius^2
    denominator <- sum(inside) - 1
    if (denominator <= 0) return(NA_real_)
    numerator <- sum(inside & component$CD8T, na.rm = TRUE) -
      as.integer(isTRUE(candidate$CD8T))
    numerator / denominator
  }

  candidates |>
    mutate(
      cd8_density = map_dbl(row_number(), density_at_center),
      center_distance = (array_col - center_x)^2 + (array_row - center_y)^2
    ) |>
    arrange(desc(cd8_density), center_distance) |>
    slice_head(n = 1)
}

excluded_low_adenoma <- pick_example_spot(
  cd8_demo_data,
  "Adenoma",
  "Low",
  prefer_cd8 = TRUE
)$spot_id

centers <- bind_rows(
  pick_example_spot(cd8_demo_data, "Adenoma", "High", prefer_cd8 = FALSE) |>
    mutate(Center = "Adenoma center"),
  pick_example_spot(cd8_demo_data, "Normal", "High", prefer_cd8 = FALSE) |>
    mutate(Center = "Normal center"),
  pick_example_spot(
    filter(cd8_demo_data, spot_id != excluded_low_adenoma),
    "Adenoma",
    "Low",
    prefer_cd8 = TRUE
  ) |>
    mutate(Center = "Adenoma center"),
  pick_example_spot(cd8_demo_data, "Normal", "Low", prefer_cd8 = TRUE) |>
    mutate(Center = "Normal center")
) |>
  mutate(
    center_id = row_number(),
    Center = factor(Center, levels = c("Adenoma center", "Normal center"))
  )

circle_data <- tidyr::crossing(
  centers |>
    select(center_id, GDF15_group, Center, Region, array_col, array_row),
  angle = seq(0, 2 * pi, length.out = 200)
) |>
  mutate(
    circle_x = array_col + radius * cos(angle),
    circle_y = array_row + radius * sin(angle)
  )

make_demo_facet <- function(data, facet_centers) {
  inside_any <- rep(FALSE, nrow(data))
  for (index in seq_len(nrow(facet_centers))) {
    same_component <- data$tissue_island == facet_centers$tissue_island[[index]]
    inside_circle <- (data$array_col - facet_centers$array_col[[index]])^2 +
      (data$array_row - facet_centers$array_row[[index]])^2 <= radius^2
    inside_any <- inside_any | (same_component & inside_circle)
  }
  data |>
    mutate(
      GDF15_group = facet_centers$GDF15_group[[1]],
      in_radius = inside_any,
      CD8_signal = if_else(CD8T, "CD8 high", "CD8 low")
    )
}

plot_data <- bind_rows(
  make_demo_facet(cd8_demo_data, filter(centers, GDF15_group == "Low")),
  make_demo_facet(cd8_demo_data, filter(centers, GDF15_group == "High"))
)

s7h <- ggplot(plot_data, aes(array_col, array_row)) +
  geom_point(aes(color = Region), size = 0.8, alpha = 0.5) +
  scale_color_manual(
    name = "Pathology",
    values = c(Normal = "#8fb8ff", Adenoma = "#ff8a8a")
  ) + ggnewscale::new_scale_color() +
  geom_point(
    data = filter(plot_data, in_radius),
    aes(color = CD8_signal),
    size = 0.8,
    alpha = 0.95
  ) +
  scale_color_manual(
    name = "CD8 signal",
    values = c("CD8 high" = "red", "CD8 low" = "grey40")
  ) +
  geom_path(
    data = circle_data,
    aes(circle_x, circle_y, group = center_id, linetype = Center),
    inherit.aes = FALSE,
    color = "black"
  ) +
  geom_point(
    data = centers,
    aes(array_col, array_row, shape = Center),
    inherit.aes = FALSE,
    color = "black",
    fill = "yellow",
    size = 3
  ) +
  scale_shape_manual(
    name = "Center",
    values = c("Adenoma center" = 21, "Normal center" = 24)
  ) +
  scale_linetype_manual(
    name = "Center",
    values = c("Adenoma center" = "solid", "Normal center" = "dashed")
  ) +
  coord_equal() +
  scale_y_reverse() +
  facet_wrap(~GDF15_group) +
  theme_classic() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(
    x = NULL,
    y = NULL,
    title = "Ring neighborhood illustration (r = 2 x spot unit)",
    subtitle = "Circles show the search area within the same tissue component"
  )
s7h
save_plot(s7h, file.path(figure_dir, "SFigure7H_CD8_neighborhood_illustration.pdf"), 8, 4)
```
