---
title: "Visium analysis workflow"
output:
  html_document:
    toc: true
    toc_float: true
params:
  execute: false
  run_preprocessing: false
  cores: 1
editor_options:
  chunk_output_type: console
---

This notebook contains only analyses reported in the Methods. Exploratory, deprecated, and figure-only code is intentionally omitted. Set `execute: true` when the inputs described in `data/README.md` are available.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = isTRUE(params$execute),
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(Seurat)
library(SeuratWrappers)
library(tidyverse)
library(Matrix)
library(qs)
library(here)

source(here("R", "preprocessing.R"))
source(here("R", "scoring.R"))
source(here("R", "Cal_senepy.R"))
source(here("R", "spatial_statistics.R"))
source(here("R", "plotting.R"))

set.seed(1)
dir.create(here("data", "derived"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results", "tables"), recursive = TRUE, showWarnings = FALSE)
```

# Visium loading, QC, and multimodal integration

RNA spots are retained when total UMI count is greater than 300 and mitochondrial RNA is below 40%. `Load10X_Spatial(filter.matrix = TRUE)` restricts loading to the Space Ranger tissue mask. ADT spots must have at least 20 detected target antibodies and less than 20% of counts in isotype controls.

```{r load-and-qc}
manifest <- read_sample_manifest(here("config", "sample_manifest.csv"))

if (isTRUE(params$run_preprocessing)) {
  visium_list <- load_visium_samples(manifest)
  visium_list <- purrr::map(visium_list, qc_rna, min_umi = 300, max_mito = 40)
  visium_list <- add_pathology_annotations(
    visium_list,
    here("data", "reference", "pathology_annotations.csv")
  )
  merged <- merge_visium_samples(visium_list)
}
```

RNA is log-normalized using 2,000 variable features, scaled, reduced by PCA, and corrected across slides with FastMNN. ADTs are DSB-normalized within slide using isotype controls, scaled, and corrected with Harmony. WNN uses 30 RNA and 15 ADT dimensions; clustering uses the SLM implementation (`algorithm = 3`).

```{r multimodal-reference}
if (isTRUE(params$run_preprocessing)) {
  integrated <- build_multimodal_reference(
    merged,
    rna_dims = 30,
    adt_dims = 15,
    resolution = 0.4,
    seed = 1
  )

  mapped_query <- map_rna_only_spots(
    reference = integrated$reference,
    query = integrated$query,
    dims = 30,
    seed = 1
  )
  combined_final <- merge_reference_and_query(integrated$reference, mapped_query)
  qs::qsave(combined_final, here("data", "derived", "visium_integrated.qs"), preset = "high")
} else {
  combined_final <- qs::qread(here("data", "derived", "visium_integrated.qs"))
}
```

# Cluster markers and annotations

```{r cluster-markers}
DefaultAssay(combined_final) <- "Spatial"
combined_final <- JoinLayers(combined_final, assay = "Spatial")
Idents(combined_final) <- factor(combined_final$combined_cluster)

cluster_markers <- FindAllMarkers(
  combined_final,
  only.pos = TRUE,
  logfc.threshold = 0,
  group.by = "combined_cluster"
) |>
  filter(p_val_adj < 0.05, !grepl("^MT-", gene)) |>
  mutate(delta_pct = pct.1 - pct.2)

write_csv(cluster_markers, here("results", "tables", "cluster_markers.csv"))

cluster_by_size <- table(combined_final$combined_cluster, combined_final$LargeSmall)
cluster_size_chisq <- chisq.test(cluster_by_size)
cluster_size_posthoc <- chisq.posthoc.test::chisq.posthoc.test(
  cluster_by_size,
  method = "fdr"
)
```

Cluster names used throughout the paper were assigned by jointly reviewing these RNA markers, normalized ADTs, and spatial localization relative to the pathology annotations. No expression-derived pathology labels are created in this notebook.

# Gene-set, senescence, cell-cycle, and developmental-potential scores

```{r irgsea}
gene_sets <- readRDS(here("data", "reference", "gene_sets.rds"))
combined_final <- run_irgsea(
  combined_final,
  gene_sets = gene_sets,
  group_by = "combined_cluster",
  assay = "Spatial"
)

combined_final <- copy_ucell_scores(
  combined_final,
  c(
    SenMayo = "SenMayo",
    IEX = "IEX",
    Stem_Signature = "Stem_Signature"
  )
)
```

```{r senepy-cell-cycle-cytotrace}
combined_final <- Calculate_senepy(combined_final, assay = "Spatial", layer = "data")
combined_final$Senepy <- combined_final$Universal_score
combined_final <- classify_senepy_outliers(
  combined_final,
  score_col = "Senepy",
  output_col = "SenePyhigh"
)
combined_final <- add_cell_cycle_scores(combined_final, assay = "Spatial")

cytotrace <- run_cytotrace2(combined_final, assay = "Spatial", ncores = params$cores)
combined_final$CytoTRACE2_Score <- cytotrace$CytoTRACE2_Score
```

# Spatial pseudotime and diffusion map

stLearn was run independently in Python/JupyterLab, not from R. Per-sample H5AD inputs must contain raw expression, `Adenoma`, `combined_cluster`, array coordinates, image coordinates, and Visium spatial metadata. See `python/README.md` for the standalone workflow. This section imports its results and calculates the diffusion map in R.

```{r import-stlearn-output}
pseudotime_files <- list.files(
  here("data", "derived", "pseudotime"),
  pattern = "[.]pseudotime[.]csv$",
  full.names = TRUE
)
pseudotime <- purrr::map_dfr(pseudotime_files, read_csv, show_col_types = FALSE)

raw_barcode <- sub("^.*_([ACGT]+-[0-9]+)$", "\\1", colnames(combined_final))
lookup <- tibble(
  seurat_barcode = colnames(combined_final),
  sample_id = combined_final$Sample_ID,
  barcode = raw_barcode
)
pseudotime <- left_join(pseudotime, lookup, by = c("sample_id", "barcode"))
combined_final$dpt_pseudotime <- pseudotime$dpt_pseudotime[
  match(colnames(combined_final), pseudotime$seurat_barcode)
]

epithelial_clusters <- c("0", "1", "2", "3", "4", "6", "8")
epithelial <- subset(
  combined_final,
  cells = colnames(combined_final)[
    combined_final$combined_cluster %in% epithelial_clusters &
      combined_final$Adenoma %in% c("Normal", "Low-grade")
  ]
)
DefaultAssay(epithelial) <- "Spatial"
epithelial[["Spatial"]] <- split(epithelial[["Spatial"]], f = epithelial$Sample_ID)
epithelial <- epithelial |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(nfeatures = 2000, verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(npcs = 20, verbose = FALSE)
epithelial <- IntegrateLayers(
  epithelial,
  method = FastMNNIntegration,
  orig.reduction = "pca",
  new.reduction = "mnn",
  verbose = FALSE
)
epithelial <- add_diffusion_map(epithelial, reduction = "mnn", dims = 1:20, sigma = 1)
qs::qsave(epithelial, here("data", "derived", "epithelial_diffusion_map.qs"))
```

# Cell2location deconvolution

Cell2location was also run independently in Python/JupyterLab. The R workflow imports only its fifth-percentile posterior abundance estimates; these estimates are not used to define CD8-positive spots or CD8 neighborhoods.

```{r import-cell2location-output}
cell2location_q05 <- read_csv(
  here("data", "derived", "cell2location", "cell2location_q05.csv"),
  show_col_types = FALSE
)
colnames(cell2location_q05)[1] <- "barcode"
rownames(cell2location_q05) <- cell2location_q05$barcode
cell2location_q05$barcode <- NULL
combined_final <- AddMetaData(
  combined_final,
  cell2location_q05[intersect(colnames(combined_final), rownames(cell2location_q05)), , drop = FALSE]
)
```

# Differential expression

All three tests use dysplastic clusters C2 and C4 where specified. Stemness cutoffs are the cohort-wide lower and upper quartiles across C2/C4; intermediate spots are removed. Senescence outliers use the cohort-wide definition above. Secreted genes are filtered after the pathologist-defined dysplastic-versus-normal test.

```{r differential-expression}
DefaultAssay(combined_final) <- "Spatial"
Idents(combined_final) <- factor(combined_final$combined_cluster)
dysplastic <- subset(combined_final, idents = c("2", "4"))

de_advanced <- FindMarkers(
  dysplastic,
  group.by = "LargeSmall",
  ident.1 = "Large",
  ident.2 = "Small",
  logfc.threshold = 0,
  min.pct = 0
)

stem_cutoffs <- quantile(dysplastic$Stem_Signature, c(0.25, 0.75), na.rm = TRUE)
dysplastic$StemGroup <- case_when(
  dysplastic$Stem_Signature <= stem_cutoffs[[1]] ~ "Low",
  dysplastic$Stem_Signature >= stem_cutoffs[[2]] ~ "High",
  TRUE ~ "Intermediate"
)
de_stemness <- FindMarkers(
  subset(dysplastic, subset = StemGroup != "Intermediate"),
  group.by = "StemGroup",
  ident.1 = "High",
  ident.2 = "Low",
  logfc.threshold = 0,
  min.pct = 0
)

senescent_epithelial <- subset(
  combined_final,
  cells = colnames(combined_final)[
    combined_final$SenePyhigh &
      combined_final$combined_cluster %in% epithelial_clusters &
      combined_final$Adenoma %in% c("Normal", "Low-grade")
  ]
)
de_senescence_region <- FindMarkers(
  senescent_epithelial,
  group.by = "Adenoma",
  ident.1 = "Low-grade",
  ident.2 = "Normal",
  logfc.threshold = 0,
  min.pct = 0
)
secreted <- read_tsv(here("data", "reference", "secreted_genes.tsv"), show_col_types = FALSE)$Gene
de_senescence_secreted <- de_senescence_region[rownames(de_senescence_region) %in% secreted, , drop = FALSE]

write_csv(rownames_to_column(de_advanced, "gene"), here("results", "tables", "de_advanced_vs_nonadvanced.csv"))
write_csv(rownames_to_column(de_stemness, "gene"), here("results", "tables", "de_stemness_high_vs_low.csv"))
write_csv(rownames_to_column(de_senescence_secreted, "gene"), here("results", "tables", "de_senescence_secreted.csv"))
```

# Bivariate local Moran's I

```{r local-moran}
moran_data <- spatial_metadata(combined_final)
moran_data$GDF15 <- extract_feature(
  combined_final,
  "GDF15",
  assay = "Spatial"
)[moran_data$barcode]
moran_data$`ADT-CD8A` <- extract_feature(
  combined_final,
  "ADT-CD8A",
  assay = "ADT_dsb"
)[moran_data$barcode]
coord_cols <- if (all(c("x", "y") %in% colnames(moran_data))) c("x", "y") else c("imagecol", "imagerow")

moran_sen_stem <- bivariate_local_moran(
  moran_data,
  x = "Senepy",
  y = "Stem_Signature",
  coord_cols = coord_cols,
  k = 6,
  permutations = 999,
  seed = 1
)
moran_gdf15_cd8 <- bivariate_local_moran(
  moran_data,
  x = "GDF15",
  y = "ADT-CD8A",
  coord_cols = coord_cols,
  k = 6,
  permutations = 999,
  seed = 1
)
write_csv(moran_sen_stem, here("results", "tables", "local_moran_senescence_stemness.csv"))
write_csv(moran_gdf15_cd8, here("results", "tables", "local_moran_gdf15_cd8.csv"))
```

# Partial spatial correlation

```{r partial-spatial-correlation}
lag_data <- spatial_metadata(
  combined_final,
  variables = c("GDF15", "Senepy", "Stem_Signature")
) |>
  filter(Adenoma == "Low-grade")

before_gdf15 <- spatial_lag_correlations(
  lag_data,
  x = "Senepy",
  y = "Stem_Signature",
  adjust_x_for = NULL,
  adjust_y_for = NULL,
  coord_cols = c("array_row", "array_col"),
  k = 6
)
after_gdf15 <- spatial_lag_correlations(
  lag_data,
  x = "Senepy",
  y = "Stem_Signature",
  adjust_x_for = "GDF15",
  adjust_y_for = "GDF15",
  coord_cols = c("array_row", "array_col"),
  k = 6
)
partial_spatial_test <- compare_before_after(
  before_gdf15,
  after_gdf15,
  permutations = 999,
  seed = 1
)
write_csv(partial_spatial_test$table, here("results", "tables", "partial_spatial_correlation.csv"))
```

# CD8 neighborhood analysis

```{r cd8-neighborhood}
cd8_data <- prepare_cd8_neighborhood_data(combined_final)
cd8_region <- cd8_density_by_source(
  cd8_data,
  comparison = "region",
  radius_multiplier = 2,
  min_sources = 5
)
cd8_gdf15 <- cd8_density_by_source(
  cd8_data,
  comparison = "gdf15",
  radius_multiplier = 2,
  min_sources = 10
)

cd8_region_test <- wilcox.test(cd8_region$Low_grade, cd8_region$Normal, paired = TRUE, exact = FALSE)
cd8_gdf15_test <- cd8_gdf15 |>
  group_by(Adenoma) |>
  summarise(
    n_samples = sum(complete.cases(GDF15_high, GDF15_low)),
    p_value = wilcox.test(GDF15_high, GDF15_low, paired = TRUE, exact = FALSE)$p.value,
    median_difference = median(GDF15_high - GDF15_low, na.rm = TRUE),
    .groups = "drop"
  )
write_csv(cd8_region, here("results", "tables", "cd8_density_region.csv"))
write_csv(cd8_gdf15, here("results", "tables", "cd8_density_gdf15_high_low.csv"))
```

# Luminal-surface distance

```{r luminal-distance}
surface_annotations <- read_csv(
  here("data", "reference", "luminal_surface_annotations.csv"),
  show_col_types = FALSE
)
surface_object <- qs::qread(here("data", "derived", "visium_semla.qs"))
surface_object <- separate_tissue_islands_semla(surface_object)
staffli_coordinates <- surface_object@tools$Staffli@meta_data |>
  as_tibble() |>
  select(barcode, pxl_col_in_fullres, pxl_row_in_fullres)
surface_data <- surface_object@meta.data |>
  rownames_to_column("barcode") |>
  select(barcode, Sample_ID, Adenoma, tissue_island, Senepy, Stem_Signature) |>
  left_join(staffli_coordinates, by = "barcode") |>
  left_join(surface_annotations, by = c("Sample_ID" = "sample_id", "barcode"))
surface_distance <- distance_to_surface(surface_data, surface_col = "surface")
surface_data <- left_join(surface_data, surface_distance, by = c("Sample_ID", "tissue_island", "barcode"))

stem_distance_bins <- summarize_distance_bins(surface_data, score = "Stem_Signature")
senescence_distance_bins <- summarize_distance_bins(surface_data, score = "Senepy")
write_csv(stem_distance_bins, here("results", "tables", "surface_distance_stemness.csv"))
write_csv(senescence_distance_bins, here("results", "tables", "surface_distance_senescence.csv"))
```

# Sample-level continuous and compositional tests

```{r sample-level-statistics}
senescence_large_small <- compare_sample_summaries(
  combined_final@meta.data,
  score = "Senepy",
  condition = "LargeSmall",
  levels = c("Small", "Large"),
  paired = FALSE
)

cluster_composition <- composition_tests(
  combined_final@meta.data,
  category = "combined_cluster",
  condition = "LargeSmall",
  sample_col = "Sample_ID",
  paired = FALSE,
  permutations = 999,
  seed = 1
)
write_csv(cluster_composition$posthoc, here("results", "tables", "cluster_composition_posthoc.csv"))
```

# Public single-cell validation

The public discovery and validation objects are restricted to tubular/tubulovillous adenomas before this step. Original published embeddings and annotations are retained in the complete objects; epithelial subsets are re-normalized and re-clustered for the subpopulation analysis.

```{r single-cell-validation}
validation_objects <- list(
  discovery = qs::qread(here("data", "reference", "validation_discovery.qs")),
  validation = qs::qread(here("data", "reference", "validation_validation.qs"))
)

recluster_epithelium <- function(object, epithelial_cells, resolution = 0.4) {
  subset(object, cells = epithelial_cells) |>
    NormalizeData(verbose = FALSE) |>
    FindVariableFeatures(nfeatures = 2000, verbose = FALSE) |>
    ScaleData(verbose = FALSE) |>
    RunPCA(npcs = 30, verbose = FALSE) |>
    FindNeighbors(dims = 1:30, verbose = FALSE) |>
    FindClusters(resolution = resolution, algorithm = 3, verbose = FALSE) |>
    RunUMAP(dims = 1:30, seed.use = 1, verbose = FALSE)
}

validation_epithelium <- purrr::map(validation_objects, function(object) {
  epithelial_cells <- rownames(object@meta.data)[object$Cell_Type == "ASC"]
  epithelial <- recluster_epithelium(object, epithelial_cells)
  epithelial <- run_irgsea(epithelial, gene_sets, group_by = "seurat_clusters", assay = "RNA")
  epithelial <- Calculate_senepy(epithelial, assay = "RNA", layer = "data")
  epithelial$Senepy <- epithelial$Universal_score
  epithelial
})
qs::qsave(validation_epithelium, here("data", "derived", "single_cell_validation.qs"))
```

```{r save-final-object}
qs::qsave(combined_final, here("data", "derived", "visium_scored.qs"), preset = "high")
writeLines(capture.output(sessionInfo()), here("results", "sessionInfo.txt"))
```
