---
title: "Visium analysis workflow"
output:
  html_document:
    toc: true
    toc_float: true
params:
  execute: false
  run_preprocessing: false
  rebuild_staffli: false
  cores: 1
editor_options:
  chunk_output_type: console
---

This notebook contains only analyses reported in the Methods. Exploratory, deprecated, and figure-only code is intentionally omitted. Set `execute: true` when the inputs described in `data/README.md` are available.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = isTRUE(params$execute),
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(Seurat)
library(SeuratWrappers)
library(tidyverse)
library(Matrix)
library(qs)
library(here)

source(here("R", "preprocessing.R"))
source(here("R", "scoring.R"))
source(here("R", "Cal_senepy.R"))
source(here("R", "spatial_statistics.R"))
source(here("R", "plotting.R"))

set.seed(1)
dir.create(here("data", "derived"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results", "tables"), recursive = TRUE, showWarnings = FALSE)
```

# Visium loading, QC, and multimodal integration

RNA spots are retained when total UMI count is greater than 300 and mitochondrial RNA is below 40%. `Load10X_Spatial(filter.matrix = TRUE)` restricts loading to the Space Ranger tissue mask. ADT spots must have at least 20 detected target antibodies and less than 20% of counts in isotype controls.

```{r load-and-qc}
manifest <- read_sample_manifest(here("config", "sample_manifest.csv"))

if (isTRUE(params$run_preprocessing)) {
  visium_list <- load_visium_samples(manifest)
  visium_list <- purrr::map(visium_list, qc_rna, min_umi = 300, max_mito = 40)
  visium_list <- add_pathology_annotations(
    visium_list,
    here("data", "reference", "pathology_annotations.csv")
  )
  merged <- merge_visium_samples(visium_list)
}
```

RNA is log-normalized using 2,000 variable features, scaled, reduced by PCA, and corrected across slides with FastMNN. ADTs are DSB-normalized within slide using isotype controls, scaled, and corrected with Harmony. WNN uses 30 RNA and 15 ADT dimensions; clustering uses the SLM implementation (`algorithm = 3`).

```{r multimodal-reference}
if (isTRUE(params$run_preprocessing)) {
  integrated <- build_multimodal_reference(
    merged,
    rna_dims = 30,
    adt_dims = 15,
    resolution = 0.4,
    seed = 1
  )

  mapped_query <- map_rna_only_spots(
    reference = integrated$reference,
    query = integrated$query,
    dims = 30,
    seed = 1
  )
  combined_final <- merge_reference_and_query(integrated$reference, mapped_query)
  qs::qsave(combined_final, here("data", "derived", "visium_integrated.qs"), preset = "high")
} else {
  combined_final <- qs::qread(here("data", "derived", "visium_integrated.qs"))
}
```

# Cluster markers and annotations

```{r cluster-markers}
DefaultAssay(combined_final) <- "Spatial"
combined_final <- JoinLayers(combined_final, assay = "Spatial")
Idents(combined_final) <- factor(combined_final$combined_cluster)

cluster_markers <- FindAllMarkers(
  combined_final,
  only.pos = TRUE,
  logfc.threshold = 0,
  group.by = "combined_cluster"
) |>
  filter(p_val_adj < 0.05, !grepl("^MT-", gene)) |>
  mutate(delta_pct = pct.1 - pct.2)

write_csv(cluster_markers, here("results", "tables", "cluster_markers.csv"))
```

Cluster names used throughout the paper were assigned by jointly reviewing these RNA markers, normalized ADTs, and spatial localization relative to the pathology annotations. No expression-derived pathology labels are created in this notebook.

# Gene-set, senescence, cell-cycle, and developmental-potential scores

```{r irgsea}
gene_sets <- readRDS(here("data", "reference", "gene_sets.rds"))
combined_final <- run_irgsea(
  combined_final,
  gene_sets = gene_sets,
  group_by = "combined_cluster",
  assay = "Spatial"
)

combined_final <- copy_ucell_scores(
  combined_final,
  c(
    SenMayo = "SenMayo",
    IEX = "IEX",
    Stem_Signature = "Stem_Signature"
  )
)
```

```{r senepy-cell-cycle-cytotrace}
combined_final <- Calculate_senepy(combined_final, assay = "Spatial", layer = "data")
combined_final$Senepy <- combined_final$Universal_score
combined_final <- classify_senepy_outliers(
  combined_final,
  score_col = "Senepy",
  output_col = "SenePyhigh"
)
combined_final <- add_cell_cycle_scores(combined_final, assay = "Spatial")

cytotrace <- run_cytotrace2(combined_final, assay = "Spatial", ncores = params$cores)
combined_final$CytoTRACE2_Score <- cytotrace$CytoTRACE2_Score
```

# Spatial pseudotime and diffusion map

stLearn was run independently in Python/JupyterLab, not from R. Per-sample H5AD inputs must contain raw expression, `Adenoma`, `combined_cluster`, array coordinates, image coordinates, and Visium spatial metadata. See `python/README.md` for the standalone workflow. This section imports its results and calculates the diffusion map in R.

```{r import-stlearn-output}
pseudotime_files <- list.files(
  here("data", "derived", "pseudotime"),
  pattern = "[.]pseudotime[.]csv$",
  full.names = TRUE
)
pseudotime <- purrr::map_dfr(pseudotime_files, read_csv, show_col_types = FALSE)

raw_barcode <- sub("^.*_([ACGT]+-[0-9]+)$", "\\1", colnames(combined_final))
lookup <- tibble(
  seurat_barcode = colnames(combined_final),
  sample_id = combined_final$Sample_ID,
  barcode = raw_barcode
)
pseudotime <- left_join(pseudotime, lookup, by = c("sample_id", "barcode"))
combined_final$dpt_pseudotime <- pseudotime$dpt_pseudotime[
  match(colnames(combined_final), pseudotime$seurat_barcode)
]

epithelial_clusters <- c("0", "1", "2", "3", "4", "6", "8")
epithelial <- subset(
  combined_final,
  cells = colnames(combined_final)[
    combined_final$combined_cluster %in% epithelial_clusters &
      combined_final$Adenoma %in% c("Normal", "Low-grade")
  ]
)
DefaultAssay(epithelial) <- "Spatial"
epithelial[["Spatial"]] <- split(epithelial[["Spatial"]], f = epithelial$Sample_ID)
epithelial <- epithelial |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(nfeatures = 2000, verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(npcs = 20, verbose = FALSE)
epithelial <- IntegrateLayers(
  epithelial,
  method = FastMNNIntegration,
  orig.reduction = "pca",
  new.reduction = "mnn",
  verbose = FALSE
)
epithelial <- add_diffusion_map(epithelial, reduction = "mnn", dims = 1:20, sigma = 1)
qs::qsave(epithelial, here("data", "derived", "epithelial_diffusion_map.qs"))
```

# Cell2location deconvolution

Cell2location was also run independently in Python/JupyterLab. The R workflow imports only its fifth-percentile posterior abundance estimates; these estimates are not used to define CD8-positive spots or CD8 neighborhoods.

```{r import-cell2location-output}
cell2location_q05 <- read_csv(
  here("data", "derived", "cell2location", "cell2location_q05.csv"),
  show_col_types = FALSE
)
colnames(cell2location_q05)[1] <- "barcode"
rownames(cell2location_q05) <- cell2location_q05$barcode
cell2location_q05$barcode <- NULL
combined_final <- AddMetaData(
  combined_final,
  cell2location_q05[intersect(colnames(combined_final), rownames(cell2location_q05)), , drop = FALSE]
)
```

# Differential expression and GDF15 candidate prioritization

This section follows the authoritative `DE analysis for GDF15` chunk in the working analysis notebook. All three contrasts retain only positive markers with a minimum detection-fraction difference of 0.10 and a log-fold-change threshold of 0.2. Advancement and stemness are tested within dysplastic clusters C2 and C4. Stemness cutoffs are the cohort-wide lower and upper quartiles across C2/C4. The senescence cutoff is the mean plus two standard deviations among non-stromal spots, followed by the pathologist-defined low-grade-versus-normal comparison. Candidate secreted genes must be positive in all three contrasts and present in the secretome reference.

```{r differential-expression-gdf15}
DefaultAssay(combined_final) <- "Spatial"
combined_final <- JoinLayers(combined_final, assay = "Spatial")
Idents(combined_final) <- factor(combined_final$combined_cluster)
dysplastic <- subset(combined_final, idents = c("2", "4"))

stem_cutoffs <- quantile(dysplastic$Stem_Signature, c(0.25, 0.75), na.rm = TRUE)
dysplastic$StemGroup <- case_when(
  dysplastic$Stem_Signature <= stem_cutoffs[[1]] ~ "Low",
  dysplastic$Stem_Signature >= stem_cutoffs[[2]] ~ "High",
  TRUE ~ "Mid"
)

de_stemness <- FindMarkers(
  dysplastic,
  group.by = "StemGroup",
  ident.1 = "High",
  ident.2 = "Low",
  min.diff.pct = 0.10,
  logfc.threshold = 0.2,
  only.pos = TRUE
)

de_advanced <- FindMarkers(
  dysplastic,
  group.by = "LargeSmall",
  ident.1 = "Large",
  ident.2 = "Small",
  min.diff.pct = 0.10,
  logfc.threshold = 0.2,
  only.pos = TRUE
)

non_stromal <- subset(combined_final, subset = Adenoma != "Stroma")
senepy_cutoff <- mean(non_stromal$Senepy) + 2 * sd(non_stromal$Senepy)
non_stromal$Senepy_pos <- case_when(
  non_stromal$Senepy <= senepy_cutoff ~ "No",
  non_stromal$Senepy > senepy_cutoff ~ "Yes"
)
senescent_non_stromal <- subset(non_stromal, subset = Senepy_pos == "Yes")

de_senescence_region <- FindMarkers(
  senescent_non_stromal,
  group.by = "Adenoma",
  ident.1 = "Low-grade",
  ident.2 = "Normal",
  min.diff.pct = 0.10,
  logfc.threshold = 0.2,
  only.pos = TRUE
)

secreted <- read_tsv(
  here("data", "reference", "secreted_genes.tsv"),
  show_col_types = FALSE
)$Gene

shared_de_genes <- Reduce(
  intersect,
  list(
    rownames(de_advanced),
    rownames(de_stemness),
    rownames(de_senescence_region)
  )
)
candidate_secreted <- shared_de_genes[shared_de_genes %in% secreted]

de_senescence_secreted <- de_senescence_region[
  rownames(de_senescence_region) %in% secreted,
  ,
  drop = FALSE
]

venn_data <- list(
  `Large vs Small` = rownames(de_advanced),
  `Stem-high vs Stem-low` = rownames(de_stemness),
  `Sene-Adenoma vs Sene-Normal` = rownames(de_senescence_secreted)
)

gdf15_de <- bind_rows(
  rownames_to_column(de_advanced, "gene") |>
    filter(gene == "GDF15") |>
    mutate(comparison = "Large vs Small"),
  rownames_to_column(de_stemness, "gene") |>
    filter(gene == "GDF15") |>
    mutate(comparison = "Stem-high vs Stem-low"),
  rownames_to_column(de_senescence_region, "gene") |>
    filter(gene == "GDF15") |>
    mutate(comparison = "Sene-Adenoma vs Sene-Normal")
)

write_csv(rownames_to_column(de_advanced, "gene"), here("results", "tables", "de_advanced_vs_nonadvanced.csv"))
write_csv(rownames_to_column(de_stemness, "gene"), here("results", "tables", "de_stemness_high_vs_low.csv"))
write_csv(rownames_to_column(de_senescence_region, "gene"), here("results", "tables", "de_senescence_low_grade_vs_normal.csv"))
write_csv(rownames_to_column(de_senescence_secreted, "gene"), here("results", "tables", "de_senescence_secreted.csv"))
write_csv(tibble(gene = candidate_secreted), here("results", "tables", "de_three_way_secreted_candidates.csv"))
write_csv(gdf15_de, here("results", "tables", "de_gdf15_three_comparisons.csv"))
saveRDS(venn_data, here("results", "tables", "de_gdf15_venn.rds"))

candidate_secreted
gdf15_de
```

# Bivariate local Moran's I

```{r local-moran}
moran_data <- spatial_metadata(combined_final)
moran_data$GDF15 <- extract_feature(
  combined_final,
  "GDF15",
  assay = "Spatial"
)[moran_data$barcode]
moran_data$`ADT-CD8A` <- extract_feature(
  combined_final,
  "ADT-CD8A",
  assay = "ADT_dsb"
)[moran_data$barcode]
coord_cols <- if (all(c("x", "y") %in% colnames(moran_data))) c("x", "y") else c("imagecol", "imagerow")

moran_sen_stem <- bivariate_local_moran(
  moran_data,
  x = "Senepy",
  y = "Stem_Signature",
  coord_cols = coord_cols,
  k = 6,
  permutations = 999,
  seed = 1
)
moran_gdf15_cd8 <- bivariate_local_moran(
  moran_data,
  x = "GDF15",
  y = "ADT-CD8A",
  coord_cols = coord_cols,
  k = 6,
  permutations = 999,
  seed = 1
)
write_csv(moran_sen_stem, here("results", "tables", "local_moran_senescence_stemness.csv"))
write_csv(moran_gdf15_cd8, here("results", "tables", "local_moran_gdf15_cd8.csv"))
```

# Partial spatial correlation

```{r partial-spatial-correlation}
lag_data <- spatial_metadata(
  combined_final,
  variables = c("GDF15", "Senepy", "Stem_Signature")
) |>
  filter(Adenoma == "Low-grade")

before_gdf15 <- spatial_lag_correlations(
  lag_data,
  x = "Senepy",
  y = "Stem_Signature",
  adjust_x_for = NULL,
  adjust_y_for = NULL,
  coord_cols = c("array_row", "array_col"),
  k = 6
)
after_gdf15 <- spatial_lag_correlations(
  lag_data,
  x = "Senepy",
  y = "Stem_Signature",
  adjust_x_for = "GDF15",
  adjust_y_for = "GDF15",
  coord_cols = c("array_row", "array_col"),
  k = 6
)
partial_spatial_test <- compare_before_after(
  before_gdf15,
  after_gdf15,
  permutations = 999,
  seed = 1
)
write_csv(partial_spatial_test$table, here("results", "tables", "partial_spatial_correlation.csv"))
```

# CD8 neighborhood analysis

This section reproduces the active analysis under `# CD8 Spatial distance` in the working `My_Visium.Rmd`. The public assay `ADT_dsb` is the within-slide DSB-normalized assay named `ADT_dsb_by_slide` in the source notebook. GDF15 and the composite CD8 gate are standardized across all ADT-complete spots within each sample before restricting to normal and low-grade epithelial clusters C0, C1, C2, C3, C4, C6, and C8. CD8-positive spots require z-scored CD3E and CD8A of at least 0.5 and the maximum z-scored CD45RA/CD45RO value of at least zero. Epithelial tissue components use a 1.6-spot-unit connection radius, and CD8 density is measured within a two-spot-unit ring in the same component. The first comparison uses at least five GDF15-high source spots per pathology region; the second uses at least ten GDF15-high or GDF15-low source spots per region and excludes pilot samples. The exploratory nearest-distance calculation is commented out in the source, so the two reported outputs reproduce the active ring-density analyses.

```{r cd8-neighborhood}
cd8_data <- prepare_cd8_neighborhood_data(
  combined_final,
  adt_assay = "ADT_dsb"
)
cd8_region <- cd8_density_by_source(
  cd8_data,
  comparison = "region",
  radius_multiplier = 2,
  min_sources = 5
)
cd8_gdf15 <- cd8_density_by_source(
  cd8_data,
  comparison = "gdf15",
  radius_multiplier = 2,
  min_sources = 10
)

cd8_region_test <- cd8_region |>
  dplyr::filter(!is.na(.data$dens_LGD), !is.na(.data$dens_Normal)) |>
  dplyr::group_by(.data$LargeSmall) |>
  dplyr::summarise(
    n_pairs = dplyr::n(),
    p_value = stats::wilcox.test(.data$dens_LGD, .data$dens_Normal, paired = TRUE)$p.value,
    median_difference = stats::median(.data$delta_dens),
    .groups = "drop"
  )
cd8_gdf15_test <- cd8_gdf15 |>
  dplyr::filter(!is.na(.data$dens_High), !is.na(.data$dens_Low)) |>
  dplyr::group_by(.data$Adenoma) |>
  dplyr::summarise(
    n_pairs = dplyr::n(),
    p_value = stats::wilcox.test(.data$dens_High, .data$dens_Low, paired = TRUE)$p.value,
    median_difference = stats::median(.data$delta_HL),
    .groups = "drop"
  )
write_csv(cd8_region, here("results", "tables", "cd8_density_region.csv"))
write_csv(cd8_gdf15, here("results", "tables", "cd8_density_gdf15_high_low.csv"))
write_csv(cd8_region_test, here("results", "tables", "cd8_density_region_tests.csv"))
write_csv(cd8_gdf15_test, here("results", "tables", "cd8_density_gdf15_high_low_tests.csv"))
```

# Adenoma radial distance from the tract surface

This section reconstructs `combined_final_semla` from `combined_final` and the Space Ranger coordinates, scale factors, and H&E images when `rebuild_staffli: true`. Only the object-construction steps required by the final analysis are retained; exploratory Staffli subsetting and viewer diagnostics are omitted. Manual annotations exported from the semla viewer use `All` for tract-surface anchors and `Island` for tissue pieces that were excluded. The distance is calculated within each semla-connected tissue component after removing `Island` annotations and singleton components. In the final plotting pipeline, surface anchors and spots within the first 100 micrometers are excluded from the summaries; remaining spots are grouped into 100-micrometer bins, and bins with fewer than five spots are removed.

```{r tract-surface-distance}
surface_annotations <- read_csv(
  here("data", "reference", "luminal_surface_annotations.csv"),
  show_col_types = FALSE
)
assert_columns(surface_annotations, c("sample_id", "barcode", "surface_label"))

if (isTRUE(params$rebuild_staffli)) {
  surface_object <- build_staffli_from_spaceranger(
    combined_final,
    manifest,
    image_height = 300
  )
  qs::qsave(
    surface_object,
    here("data", "derived", "visium_semla.qs"),
    preset = "high"
  )
} else {
  surface_object <- qs::qread(here("data", "derived", "visium_semla.qs"))
}

surface_object <- separate_tissue_islands_semla(surface_object)
staffli_coordinates <- surface_object@tools$Staffli@meta_data |>
  as_tibble() |>
  select(barcode, pxl_col_in_fullres, pxl_row_in_fullres)

surface_annotations <- surface_annotations |>
  transmute(
    sample_id = as.character(sample_id),
    raw_barcode = barcode,
    surface_label = case_when(
      toupper(surface_label) == "ALL" ~ "All",
      toupper(surface_label) == "ISLAND" ~ "Island",
      TRUE ~ as.character(surface_label)
    )
  )
surface_data <- surface_object@meta.data |>
  rownames_to_column("barcode") |>
  select(barcode, Sample_ID, Adenoma, tissue_island, Senepy, Stem_Signature) |>
  mutate(raw_barcode = str_extract(barcode, "[ACGT]{16}-[0-9]+")) |>
  left_join(staffli_coordinates, by = "barcode") |>
  left_join(
    surface_annotations,
    by = c("Sample_ID" = "sample_id", "raw_barcode")
  ) |>
  mutate(
    surface_label = replace_na(surface_label, "Other"),
    Adenoma = recode(as.character(Adenoma), `Low-grade` = "Adenoma", `2` = "Normal")
  ) |>
  filter(
    surface_label != "Island",
    !tolower(tissue_island) %in% c("singleton", "singletons")
  )

surface_distance <- distance_to_surface(
  surface_data,
  surface_col = "surface_label",
  surface_value = "All"
)
surface_data <- left_join(
  surface_data,
  surface_distance,
  by = c("Sample_ID", "tissue_island", "barcode")
)

write_csv(surface_data, here("results", "tables", "surface_distance_spots.csv"))
surface_analysis_data <- filter(surface_data, Adenoma != "Stroma")
stem_distance_bins <- summarize_distance_bins(surface_analysis_data, score = "Stem_Signature")
senescence_distance_bins <- summarize_distance_bins(surface_analysis_data, score = "Senepy")
write_csv(stem_distance_bins, here("results", "tables", "surface_distance_stemness.csv"))
write_csv(senescence_distance_bins, here("results", "tables", "surface_distance_senescence.csv"))
```

# Continuous and categorical-proportion tests

All continuous-value comparisons use `compare_score_clusters_paired()`. Spot-level conditions are summarized within each sample and compared with a paired Wilcoxon signed-rank test; sample-level conditions are summarized per sample and compared with an unpaired Wilcoxon rank-sum test. The reported effect is the median for `level_a` minus the median for `level_b`.

The Senepy comparisons correspond to the sample-level cluster summary plotted in `Supple_plotting.Rmd`. Pseudotime is compared between paired Low-grade and Normal regions, and between Large and Small adenomas after restricting the latter analysis to Low-grade spots. Every continuous comparison uses per-sample medians and requires at least 30 spots in an analyzed sample/condition stratum. Bonferroni-adjusted p-values are reported for the source notebook's two primary Senepy contrasts, cluster 2 versus 3 and cluster 4 versus 3.

All categorical-proportion comparisons use `run_comp_tests()`. The function forms a complete category-proportion table for every sample/condition stratum, applies the Hellinger transformation before PERMANOVA and its dispersion check, and performs category-wise Wilcoxon tests with Benjamini-Hochberg correction. Pathology is a spot-level condition, so Normal and Low-grade regions are paired within samples that contain both. Adenoma post-hoc effects are Low-grade minus Normal. Adenoma size is a sample-level condition, so Small and Large samples are compared as independent groups. The same helper can be used for other categorical outcomes, including Moran's I classifications, by changing `category_col`.

```{r sample-level-statistics}
senepy_c2_vs_c3 <- compare_score_clusters_paired(
  combined_final@meta.data,
  score_col = "Senepy",
  sample_col = "Sample_ID",
  condition_col = "combined_cluster",
  level_a = "2",
  level_b = "3",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

senepy_c4_vs_c3 <- compare_score_clusters_paired(
  combined_final@meta.data,
  score_col = "Senepy",
  sample_col = "Sample_ID",
  condition_col = "combined_cluster",
  level_a = "4",
  level_b = "3",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

senepy_c4_vs_c2 <- compare_score_clusters_paired(
  combined_final@meta.data,
  score_col = "Senepy",
  sample_col = "Sample_ID",
  condition_col = "combined_cluster",
  level_a = "4",
  level_b = "2",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

epithelial_metadata <- combined_final@meta.data |>
  filter(Adenoma %in% c("Normal", "Low-grade"))
pseudotime_lowgrade_vs_normal <- compare_score_clusters_paired(
  epithelial_metadata,
  score_col = "dpt_pseudotime",
  sample_col = "Sample_ID",
  condition_col = "Adenoma",
  level_a = "Low-grade",
  level_b = "Normal",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

low_grade_metadata <- combined_final@meta.data |>
  filter(Adenoma == "Low-grade")
pseudotime_large_vs_small_lowgrade <- compare_score_clusters_paired(
  low_grade_metadata,
  score_col = "dpt_pseudotime",
  sample_col = "Sample_ID",
  condition_col = "LargeSmall",
  level_a = "Large",
  level_b = "Small",
  condition_level = "sample",
  agg = "median",
  min_spots = 30
)

continuous_score_tests <- list(
  senepy_c2_vs_c3 = senepy_c2_vs_c3,
  senepy_c4_vs_c3 = senepy_c4_vs_c3,
  senepy_c4_vs_c2 = senepy_c4_vs_c2,
  pseudotime_lowgrade_vs_normal = pseudotime_lowgrade_vs_normal,
  pseudotime_large_vs_small_lowgrade = pseudotime_large_vs_small_lowgrade
)

senepy_primary_p_adjusted <- p.adjust(
  c(senepy_c2_vs_c3$test$p.value, senepy_c4_vs_c3$test$p.value),
  method = "bonferroni"
)

continuous_test_summary <- tibble(
  comparison = names(continuous_score_tests),
  score = c("Senepy", "Senepy", "Senepy", "dpt_pseudotime", "dpt_pseudotime"),
  condition = c(
    "combined_cluster", "combined_cluster", "combined_cluster", "Adenoma", "LargeSmall"
  ),
  level_a = c("2", "4", "4", "Low-grade", "Large"),
  level_b = c("3", "3", "2", "Normal", "Small"),
  mode = vapply(continuous_score_tests, `[[`, character(1), "mode"),
  n_a = vapply(
    continuous_score_tests,
    function(x) if (is.null(x$n_a)) x$n_samples else x$n_a,
    numeric(1)
  ),
  n_b = vapply(
    continuous_score_tests,
    function(x) if (is.null(x$n_b)) x$n_samples else x$n_b,
    numeric(1)
  ),
  median_a = vapply(continuous_score_tests, function(x) x$ave_median[[1]], numeric(1)),
  median_b = vapply(continuous_score_tests, function(x) x$ave_median[[2]], numeric(1)),
  median_difference_a_minus_b = vapply(continuous_score_tests, `[[`, numeric(1), "diff_median"),
  statistic = vapply(continuous_score_tests, function(x) unname(x$test$statistic), numeric(1)),
  p_value = vapply(continuous_score_tests, function(x) x$test$p.value, numeric(1)),
  p_adjusted_bonferroni = c(senepy_primary_p_adjusted, NA_real_, NA_real_, NA_real_)
)

write_csv(continuous_test_summary, here("results", "tables", "continuous_score_tests.csv"))
saveRDS(continuous_score_tests, here("results", "tables", "continuous_score_tests.rds"))
continuous_test_summary

test_adenoma <- run_comp_tests(
  combined_final@meta.data,
  sample_col = "Sample_ID",
  condition_col = "Adenoma",
  category_col = "combined_cluster",
  condition_level = "spot",
  condition_keep = c("Normal", "Low-grade"),
  n_perm = 999,
  seed = 1,
  paired_only = TRUE
)

test_largesmall <- run_comp_tests(
  combined_final@meta.data,
  sample_col = "Sample_ID",
  condition_col = "LargeSmall",
  category_col = "combined_cluster",
  condition_level = "sample",
  condition_keep = c("Small", "Large"),
  n_perm = 999,
  seed = 1,
  paired_only = FALSE
)

write_composition_outputs <- function(result, comparison_name) {
  output_prefix <- here("results", "tables", paste0("cluster_composition_", comparison_name))
  write_csv(result$tab, paste0(output_prefix, "_proportions.csv"))
  write_csv(result$posthoc, paste0(output_prefix, "_posthoc.csv"))
  write_csv(
    rownames_to_column(as.data.frame(result$permanova), "term"),
    paste0(output_prefix, "_permanova.csv")
  )
  writeLines(
    capture.output(result$betadisper_permutest),
    paste0(output_prefix, "_betadisper.txt")
  )
  invisible(result)
}

write_composition_outputs(test_adenoma, "adenoma")
write_composition_outputs(test_largesmall, "largesmall")
saveRDS(
  list(adenoma = test_adenoma, largesmall = test_largesmall),
  here("results", "tables", "cluster_composition_tests.rds")
)

test_adenoma$permanova
test_adenoma$betadisper_permutest
test_adenoma$posthoc
test_largesmall$permanova
test_largesmall$betadisper_permutest
test_largesmall$posthoc
```

# Public single-cell validation

The public discovery and validation objects are restricted to tubular/tubulovillous adenomas before this step. Original published embeddings and annotations are retained in the complete objects; epithelial subsets are re-normalized and re-clustered for the subpopulation analysis.

```{r single-cell-validation}
validation_objects <- list(
  discovery = qs::qread(here("data", "reference", "validation_discovery.qs")),
  validation = qs::qread(here("data", "reference", "validation_validation.qs"))
)

recluster_epithelium <- function(object, epithelial_cells, resolution = 0.4) {
  subset(object, cells = epithelial_cells) |>
    NormalizeData(verbose = FALSE) |>
    FindVariableFeatures(nfeatures = 2000, verbose = FALSE) |>
    ScaleData(verbose = FALSE) |>
    RunPCA(npcs = 30, verbose = FALSE) |>
    FindNeighbors(dims = 1:30, verbose = FALSE) |>
    FindClusters(resolution = resolution, algorithm = 3, verbose = FALSE) |>
    RunUMAP(dims = 1:30, seed.use = 1, verbose = FALSE)
}

validation_epithelium <- purrr::map(validation_objects, function(object) {
  epithelial_cells <- rownames(object@meta.data)[object$Cell_Type == "ASC"]
  epithelial <- recluster_epithelium(object, epithelial_cells)
  epithelial <- run_irgsea(epithelial, gene_sets, group_by = "seurat_clusters", assay = "RNA")
  epithelial <- Calculate_senepy(epithelial, assay = "RNA", layer = "data")
  epithelial$Senepy <- epithelial$Universal_score
  epithelial
})
qs::qsave(validation_epithelium, here("data", "derived", "single_cell_validation.qs"))
```

```{r save-final-object}
qs::qsave(combined_final, here("data", "derived", "visium_scored.qs"), preset = "high")
writeLines(capture.output(sessionInfo()), here("results", "sessionInfo.txt"))
```
