---
title: "Visium analysis workflow"
output:
  html_document:
    toc: true
    toc_float: true
params:
  execute: false
  run_preprocessing: false
  build_scrna_reference: false
  rebuild_staffli: false
  cores: 24
editor_options:
  chunk_output_type: inline
---

This notebook contains the analyses reported in the Methods. Set `execute: true` when the inputs described in `data/README.md` are available.

```{r setup, include=FALSE}

if (!exists("params", inherits = FALSE)) {
  params <- list(
    execute = TRUE,
    run_preprocessing = FALSE,
    build_scrna_reference = FALSE,
    rebuild_staffli = FALSE,
    cores = 24
  )
}

knitr::opts_chunk$set(
  echo = TRUE,
  eval = isTRUE(params$execute),
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(Seurat)
library(SeuratWrappers)
library(tidyverse)
library(Matrix)
library(qs)
library(here)

source(here("R", "preprocessing.R"))
source(here("R", "scoring.R"))
source(here("R", "Cal_senepy.R"))
source(here("R", "spatial_statistics.R"))
source(here("R", "scrna_reference.R"))
source(here("R", "plotting.R"))

set.seed(1)
dir.create(here("data", "derived"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results", "tables"), recursive = TRUE, showWarnings = FALSE)
```

# Visium loading, QC, and multimodal integration

RNA spots are retained when total UMI count is greater than 300 and mitochondrial RNA is below 40%. `Load10X_Spatial(filter.matrix = TRUE)` restricts loading to the Space Ranger tissue mask. ADT spots must have at least 20 detected target antibodies and less than 20% of counts in isotype controls.

```{r load-and-qc}
manifest <- read_sample_manifest(here("config", "sample_manifest.csv"))

if (isTRUE(params$run_preprocessing)) {
  visium_list <- load_visium_samples(manifest)
  visium_list <- purrr::map(visium_list, qc_rna, min_umi = 300, max_mito = 40)
  visium_list <- add_pathology_annotations(
    visium_list,
    here("data", "reference", "pathology_annotations.csv")
  )
  merged <- merge_visium_samples(visium_list)
}
```

RNA is log-normalized using 2,000 variable features, scaled, reduced by PCA, and corrected across slides with FastMNN. ADTs are DSB-normalized within slide using isotype controls, scaled, and corrected with Harmony. WNN uses 30 RNA and 15 ADT dimensions; clustering uses the SLM implementation (`algorithm = 3`).

```{r multimodal-reference}
if (isTRUE(params$run_preprocessing)) {
  integrated <- build_multimodal_reference(
    merged,
    rna_dims = 30,
    adt_dims = 15,
    resolution = 0.4,
    seed = 1
  )

  mapped_query <- map_rna_only_spots(
    reference = integrated$reference,
    query = integrated$query,
    dims = 30,
    seed = 1
  )
  combined_final <- merge_reference_and_query(integrated$reference, mapped_query)
  qs::qsave(combined_final, here("data", "derived", "visium_integrated.qs"), preset = "high")
} else {
  combined_final <- read_analysis_archive(
    here("data", "derived", "visium_integrated.qs")
  )
  combined_final <- prepare_integrated_checkpoint(combined_final)
}
```

# Cluster markers and annotations

```{r cluster-markers}
DefaultAssay(combined_final) <- "Spatial"
combined_final <- JoinLayers(combined_final, assay = "Spatial")
Idents(combined_final) <- factor(combined_final$combined_cluster)

cluster_markers <- FindAllMarkers(
  combined_final,
  only.pos = TRUE,
  logfc.threshold = 0,
  group.by = "combined_cluster"
) |>
  filter(p_val_adj < 0.05, !grepl("^MT-", gene)) |>
  mutate(delta_pct = pct.1 - pct.2) |> group_by(cluster) |> slice_head(n = 100)

write_csv(cluster_markers, here("results", "tables", "cluster_markers.csv"))
```

Cluster names used throughout the paper were assigned by jointly reviewing these RNA markers, normalized ADTs, and spatial localization relative to the pathology annotations. No expression-derived pathology labels are created in this notebook.

# Gene-set, senescence, cell-cycle, and developmental-potential scores

CytoTRACE2 is run only on the predominantly epithelial clusters C0, C1, C2, C3, C4, and C6. The joined `Spatial` assay is copied to a conventional `RNA` assay, the object is reduced to that assay plus the existing WNN UMAP, and raw counts are supplied to CytoTRACE2. Scores are aligned back to the full object by barcode; excluded clusters remain `NA`.

```{r irgsea}
gene_sets <- readRDS(here("data", "reference", "gene_sets.rds"))
combined_final <- run_irgsea(
  combined_final,
  gene_sets = gene_sets,
  group_by = "combined_cluster",
  assay = "Spatial"
)

combined_final <- copy_ucell_scores(
  combined_final,
  c(
    SenMayo = "SAUL_SEN_MAYO",
    IEX = "IEX_score",
    Stem_Signature = "Stem_Signature",
    E2F_TARGETS = "HALLMARK_E2F_TARGETS",
    MYC_TARGETS = "HALLMARK_MYC_TARGETS_V1",
    G2M_CHECKPOINT = "HALLMARK_G2M_CHECKPOINT",
    TGF_BETA_SIGNALING = "HALLMARK_TGF_BETA_SIGNALING",
    WNT_BETA_CATENIN_SIGNALING = "HALLMARK_WNT_BETA_CATENIN_SIGNALING",
    P53_PATHWAY = "HALLMARK_P53_PATHWAY",
    INFLAMMATORY_RESPONSE = "HALLMARK_INFLAMMATORY_RESPONSE"
  )
)

result.dge <- irGSEA.integrate(object = combined_final, 
                               group.by = "combined_cluster",
                               metadata = NULL, col.name = NULL,
                               method = c("AUCell","UCell","singscore",
                                           "JASMINE"))
saveRDS(result.dge,here("data", "derived", "result.dge.rds"))
```

```{r senepy-cell-cycle-cytotrace}
combined_final <- Calculate_senepy(combined_final, assay = "Spatial", layer = "data")
combined_final$Senepy <- combined_final$Universal_score
combined_final <- classify_senepy_outliers(
  combined_final,
  score_col = "Senepy",
  output_col = "SenePyhigh"
)
combined_final <- add_cell_cycle_scores(combined_final, assay = "Spatial")

cytotrace_input <- prepare_cytotrace2_input(
  combined_final,
  clusters = c("0", "1", "2", "3", "4", "6"),
  cluster_col = "combined_cluster",
  source_assay = "Spatial",
  output_assay = "RNA",
  reduction = "wnn.umap"
)
cytotrace <- run_cytotrace2(cytotrace_input, assay = "RNA", ncores = params$cores)
cytotrace_scores <- cytotrace@meta.data$CytoTRACE2_Score
names(cytotrace_scores) <- rownames(cytotrace@meta.data)
combined_final$CytoTRACE2_Score <- cytotrace_scores[colnames(combined_final)]
```

# Spatial pseudotime and diffusion map

stLearn was run independently in Python/JupyterLab, not from R. Per-sample H5AD inputs must contain raw expression, `Adenoma`, `combined_cluster`, array coordinates, image coordinates, and Visium spatial metadata. See `python/README.md` for the standalone workflow. This section imports the pseudotime results. For the epithelial diffusion map, clusters 0, 1, 2, 3, 4, 6, and 8 are retained without an additional pathology-region filter; RNA layers are split by sample and normalized; a 50-component PCA is followed by FastMNN; and diffusion mapping uses MNN dimensions 1–20 with `sigma = 1`. The displayed `DC1` and `DC2` are diffusion eigenvectors 1 and 3, respectively. The sign of `DC1` is oriented to correlate positively with imported `dpt_pseudotime`.

```{r import-stlearn-output}
pseudotime_files <- list.files(
  here("data", "derived", "pseudotime"),
  pattern = "[.]pseudotime[.]csv$",
  full.names = TRUE
)
pseudotime <- purrr::map_dfr(pseudotime_files, read_csv, show_col_types = FALSE)

raw_barcode <- sub("^.*_([ACGT]+-[0-9]+)$", "\\1", colnames(combined_final))
lookup <- tibble(
  seurat_barcode = colnames(combined_final),
  sample_id = combined_final$Sample_ID,
  barcode = raw_barcode
)
pseudotime <- left_join(pseudotime, lookup, by = c("sample_id", "barcode"))
combined_final$dpt_pseudotime <- pseudotime$dpt_pseudotime[
  match(colnames(combined_final), pseudotime$seurat_barcode)
]

epithelial_clusters <- c("0", "1", "2", "3", "4", "6", "8")
epithelial <- subset(
  combined_final,
  cells = colnames(combined_final)[
    combined_final$combined_cluster %in% epithelial_clusters
  ]
)
DefaultAssay(epithelial) <- "Spatial"
epithelial[["Spatial"]] <- split(epithelial[["Spatial"]], f = epithelial$Sample_ID)
epithelial <- epithelial |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(nfeatures = 2000, verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(npcs = 50, verbose = FALSE)
epithelial <- IntegrateLayers(
  epithelial,
  method = FastMNNIntegration,
  orig.reduction = "pca",
  new.reduction = "mnn",
  verbose = FALSE
)
epithelial <- add_diffusion_map(
  epithelial,
  reduction = "mnn",
  dims = 1:20,
  sigma = 1,
  components = c(1, 3),
  orient_by = "dpt_pseudotime",
  orient_component = 1
)
```

# Public scRNA-seq reference for Cell2location

The Cell2location reference is derived from the integrated public `marteau_Cancer_Cell_2026` object (https://crc.icbi.at/), not from the two Lau adenoma validation cohorts analyzed below. Missing primary annotations are filled from `cell_type_fine`; fibroblast, endothelial, cycling T-cell, cDC, and monocyte labels are consolidated; the 12 explicitly excluded rare or unwanted annotations are removed; and dysplastic labels are replaced by the curated `my_annotationV2` labels. `Dysplastic_Normal` cells are then excluded and `make.names()` is applied to the final labels used by Cell2location.

The realized reference contains 52,871 cells, 24,593 genes, 66 samples, and 28 final annotations. `validate_marteau_reference()` checks all of these values and the per-annotation cell counts against the H5AD used in the analysis. The H5AD export places raw RNA counts in `X` and retains `nCount_RNA`, `nFeature_RNA`, the original barcode index, `sample_id`, `my_annotation`, and `study_id` in `obs`. Because the export densifies the full count matrix and requires substantial memory, it runs only when `build_scrna_reference: true`.

```{r build-scrna-reference}
if (isTRUE(params$build_scrna_reference)) {
  marteau_ta <- qs::qread(here("data", "reference", "marteau_TA_updated.qs"))
  scrna_reference <- curate_marteau_reference(marteau_ta)

  if (!"Stem_Signature_UCell" %in% colnames(scrna_reference@meta.data)) {
    scrna_reference <- UCell::AddModuleScore_UCell(
      scrna_reference,
      features = list(Stem_Signature = gene_sets$Stem_Signature),
      assay = "RNA",
      BPPARAM = BiocParallel::MulticoreParam(workers = params$cores)
    )
  }
  if (!all(c("epithelial_cell_score_0", "Universal_score") %in%
      colnames(scrna_reference@meta.data))) {
    scrna_reference <- Calculate_senepy(
      scrna_reference,
      assay = "RNA",
      layer = "data"
    )
  }

  reference_counts <- validate_marteau_reference(scrna_reference)

  write_csv(
    reference_counts,
    here("results", "tables", "scrna_reference_cell_counts.csv")
  )
  qs::qsave(
    scrna_reference,
    here("data", "derived", "marteau_cell2location_reference.qs"),
    preset = "high"
  )
  export_marteau_reference_h5ad(
    scrna_reference,
    here("data", "reference", "scrna_reference.h5ad")
  )
}
```

# Cell2location deconvolution

Cell2location was also run independently in Python/JupyterLab. The R workflow imports only its fifth-percentile posterior abundance estimates; these estimates are not used to define CD8-positive spots or CD8 neighborhoods.

```{r import-cell2location-output}
cell2location_q05 <- read_csv(
  here("data", "derived", "cell2location", "cell2location_q05.csv"),
  show_col_types = FALSE
) |>
  as.data.frame()
colnames(cell2location_q05)[1] <- "barcode"
rownames(cell2location_q05) <- cell2location_q05$barcode
cell2location_q05$barcode <- NULL
cell2location_q05 <- cell2location_q05 |>
  mutate(
    Dysplastic_Epi = Dysplastic_Absorptive_like + Dysplastic_Proliferating +
      Dysplastic_Senescent_like + Dysplastic_Stem_like +
      Dysplastic_Tuft_like + Dysplastic_Paneth_like,
    Normal_Epi = Normal.Goblet + Normal.Crypt.cell + Normal.TA +
      Normal.Colonocyte + Normal.Colonocyte.BEST4 + Tuft,
    CD8_T_cells = CD8.T,
    CD4_T_cells = CD4.T,
    B_cell = B.cell.activated + GC.B.cell,
    Plasma_cell = Plasma.IgA + Plasma.IgG,
    T_cells = CD8_T_cells + CD4_T_cells
  )
combined_final <- AddMetaData(
  combined_final,
  cell2location_q05[intersect(colnames(combined_final), rownames(cell2location_q05)), , drop = FALSE]
)
```

# Differential expression and GDF15 candidate prioritization

All three contrasts retain only positive markers with a minimum detection-fraction difference of 0.10 and a log-fold-change threshold of 0.2. Advancement and stemness are tested within dysplastic clusters C2 and C4. Stemness cutoffs are the cohort-wide lower and upper quartiles across C2/C4. The senescence cutoff is the mean plus two standard deviations among non-stromal spots, followed by the pathologist-defined low-grade-versus-normal comparison. Candidate secreted genes must be positive in all three contrasts and present in the secretome reference.

```{r differential-expression-gdf15}
DefaultAssay(combined_final) <- "Spatial"
combined_final <- JoinLayers(combined_final, assay = "Spatial")
Idents(combined_final) <- factor(combined_final$combined_cluster)
dysplastic <- subset(combined_final, idents = c("2", "4"))

stem_cutoffs <- quantile(dysplastic$Stem_Signature, c(0.25, 0.75), na.rm = TRUE)
dysplastic$StemGroup <- case_when(
  dysplastic$Stem_Signature <= stem_cutoffs[[1]] ~ "Low",
  dysplastic$Stem_Signature >= stem_cutoffs[[2]] ~ "High",
  TRUE ~ "Mid"
)

de_stemness <- FindMarkers(
  dysplastic,
  group.by = "StemGroup",
  ident.1 = "High",
  ident.2 = "Low",
  min.diff.pct = 0.10,
  logfc.threshold = 0.2,
  only.pos = TRUE
)

de_advanced <- FindMarkers(
  dysplastic,
  group.by = "LargeSmall",
  ident.1 = "Large",
  ident.2 = "Small",
  min.diff.pct = 0.10,
  logfc.threshold = 0.2,
  only.pos = TRUE
)

non_stromal <- subset(combined_final, subset = Adenoma != "Stroma")
senepy_cutoff <- mean(non_stromal$Senepy) + 2 * sd(non_stromal$Senepy)
non_stromal$Senepy_pos <- case_when(
  non_stromal$Senepy <= senepy_cutoff ~ "No",
  non_stromal$Senepy > senepy_cutoff ~ "Yes"
)
senescent_non_stromal <- subset(non_stromal, subset = Senepy_pos == "Yes")

de_senescence_region <- FindMarkers(
  senescent_non_stromal,
  group.by = "Adenoma",
  ident.1 = "Low-grade",
  ident.2 = "Normal",
  min.diff.pct = 0.10,
  logfc.threshold = 0.2,
  only.pos = TRUE
)

secreted <- read_tsv(
  here("data", "reference", "secreted_genes.tsv"),
  show_col_types = FALSE
)$Gene

shared_de_genes <- Reduce(
  intersect,
  list(
    rownames(de_advanced),
    rownames(de_stemness),
    rownames(de_senescence_region)
  )
)
candidate_secreted <- shared_de_genes[shared_de_genes %in% secreted]

de_senescence_secreted <- de_senescence_region[
  rownames(de_senescence_region) %in% secreted,
  ,
  drop = FALSE
]

venn_data <- list(
  `Large vs Small` = rownames(de_advanced),
  `Stem-high vs Stem-low` = rownames(de_stemness),
  `Sene-Adenoma vs Sene-Normal` = rownames(de_senescence_secreted)
)

gdf15_de <- bind_rows(
  rownames_to_column(de_advanced, "gene") |>
    filter(gene == "GDF15") |>
    mutate(comparison = "Large vs Small"),
  rownames_to_column(de_stemness, "gene") |>
    filter(gene == "GDF15") |>
    mutate(comparison = "Stem-high vs Stem-low"),
  rownames_to_column(de_senescence_region, "gene") |>
    filter(gene == "GDF15") |>
    mutate(comparison = "Sene-Adenoma vs Sene-Normal")
)

write_csv(rownames_to_column(de_advanced, "gene"), here("results", "tables", "de_advanced_vs_nonadvanced.csv"))
write_csv(rownames_to_column(de_stemness, "gene"), here("results", "tables", "de_stemness_high_vs_low.csv"))
write_csv(rownames_to_column(de_senescence_region, "gene"), here("results", "tables", "de_senescence_low_grade_vs_normal.csv"))
write_csv(rownames_to_column(de_senescence_secreted, "gene"), here("results", "tables", "de_senescence_secreted.csv"))
write_csv(tibble(gene = candidate_secreted), here("results", "tables", "de_three_way_secreted_candidates.csv"))
write_csv(gdf15_de, here("results", "tables", "de_gdf15_three_comparisons.csv"))
saveRDS(venn_data, here("results", "tables", "de_gdf15_venn.rds"))

candidate_secreted
gdf15_de
```

# Bivariate local Moran's I

Five bivariate local Moran analyses are calculated with six-nearest-neighbor spatial weights: GDF15 with Senepy, ADT CD8A, Stem Signature, or Cell2location CD8 T-cell abundance, and Senepy with Stem Signature. The ADT assay is unavailable for some spots; those spots remain missing for the GDF15-ADT analysis. Negative DSB-normalized ADT-CD8A values are clipped to zero before calculating that pair. Each function call returns a metadata data frame, and the five data frames are added to the complete Visium object together after barcode alignment.

```{r local-moran}
moran_specifications <- list(
  GDF15_Senepy = list(
    target1 = "GDF15",
    target2 = "Senepy",
    assay1 = "Spatial",
    assay2 = NULL,
    clip_neg_to_zero = character(0),
    output_file = "local_moran_gdf15_senepy.csv"
  ),
  GDF15_ADT_CD8A = list(
    target1 = "GDF15",
    target2 = "ADT-CD8A",
    assay1 = "Spatial",
    assay2 = "ADT_dsb_by_slide",
    clip_neg_to_zero = "ADT-CD8A",
    output_file = "local_moran_gdf15_cd8.csv"
  ),
  GDF15_Stem_Signature = list(
    target1 = "GDF15",
    target2 = "Stem_Signature",
    assay1 = "Spatial",
    assay2 = NULL,
    clip_neg_to_zero = character(0),
    output_file = "local_moran_gdf15_stem_signature.csv"
  ),
  Senepy_Stem_Signature = list(
    target1 = "Senepy",
    target2 = "Stem_Signature",
    assay1 = NULL,
    assay2 = NULL,
    clip_neg_to_zero = character(0),
    output_file = "local_moran_senescence_stemness.csv"
  ),
  GDF15_CD8_T = list(
    target1 = "GDF15",
    target2 = "CD8.T",
    assay1 = "Spatial",
    assay2 = NULL,
    clip_neg_to_zero = character(0),
    output_file = "local_moran_gdf15_cd8_t.csv"
  )
)

moran_results <- purrr::map(moran_specifications, function(specification) {
  bivar_local_moran_seurat(
    combined_final,
    target1 = specification$target1,
    target2 = specification$target2,
    assay1 = specification$assay1,
    assay2 = specification$assay2,
    k = 6,
    clip_neg_to_zero = specification$clip_neg_to_zero,
    add_to_object = FALSE,
    verbose = FALSE
  )
})

moran_metadata <- do.call(cbind, unname(moran_results))
stopifnot(
  identical(rownames(moran_metadata), colnames(combined_final)),
  !anyDuplicated(colnames(moran_metadata))
)
combined_final <- AddMetaData(combined_final, moran_metadata)

purrr::walk2(moran_results, moran_specifications, function(result, specification) {
  write_csv(
    rownames_to_column(result, "barcode"),
    here("results", "tables", specification$output_file)
  )
})
```

The composition tests are applied to all five Moran classifications in two ways. For the sample-level size comparison, the analysis retains low-grade spots with a non-missing classification other than `Not sig`, removes samples with 10 or fewer retained spots, and compares Small and Large adenomas as independent groups. For the regional comparison, it retains Normal and Low-grade spots with an informative Moran classification and performs a paired, within-sample comparison of the two pathology regions.

```{r local-moran-composition-tests}
moran_class_columns <- paste0("Cluster_Moran_", names(moran_specifications))
stopifnot(all(moran_class_columns %in% colnames(combined_final@meta.data)))

moran_largesmall_tests <- stats::setNames(vector("list", length(moran_class_columns)), moran_class_columns)
moran_adenoma_tests <- stats::setNames(vector("list", length(moran_class_columns)), moran_class_columns)

write_moran_composition_outputs <- function(test, output_stem) {
  write_csv(test$tab, paste0(output_stem, "_proportions.csv"))
  write_csv(test$posthoc, paste0(output_stem, "_posthoc.csv"))
  write_csv(
    rownames_to_column(as.data.frame(test$permanova), "term"),
    paste0(output_stem, "_permanova.csv")
  )
  writeLines(
    capture.output(test$betadisper_permutest),
    paste0(output_stem, "_betadisper.txt")
  )
  invisible(test)
}

for (moran_column in moran_class_columns) {
  moran_largesmall_data <- combined_final@meta.data |>
    tibble::rownames_to_column("barcode") |>
    dplyr::filter(
      .data$Adenoma == "Low-grade",
      !is.na(.data[[moran_column]]),
      .data[[moran_column]] != "Not sig"
    )

  eligible_samples <- moran_largesmall_data |>
    dplyr::count(.data$Sample_ID, name = "n_spots") |>
    dplyr::filter(.data$n_spots > 10) |>
    dplyr::pull(.data$Sample_ID)

  moran_largesmall_data <- moran_largesmall_data |>
    dplyr::filter(.data$Sample_ID %in% eligible_samples)

  largesmall_test <- run_comp_tests(
    moran_largesmall_data,
    sample_col = "Sample_ID",
    condition_col = "LargeSmall",
    category_col = moran_column,
    condition_level = "sample",
    condition_keep = c("Small", "Large"),
    n_perm = 999,
    seed = 1,
    paired_only = FALSE
  )

  moran_largesmall_tests[[moran_column]] <- list(
    proportions = largesmall_test$tab,
    permanova = largesmall_test$permanova,
    betadisper = largesmall_test$betadisper_permutest,
    posthoc = largesmall_test$posthoc
  )

  largesmall_output_stem <- here(
    "results",
    "tables",
    paste0("local_moran_largesmall_", tolower(sub("^Cluster_Moran_", "", moran_column)))
  )
  write_moran_composition_outputs(largesmall_test, largesmall_output_stem)

  moran_adenoma_data <- combined_final@meta.data |>
    tibble::rownames_to_column("barcode") |>
    dplyr::filter(
      .data$Adenoma %in% c("Normal", "Low-grade"),
      !is.na(.data[[moran_column]]),
      .data[[moran_column]] != "Not sig"
    )

  adenoma_test <- run_comp_tests(
    moran_adenoma_data,
    sample_col = "Sample_ID",
    condition_col = "Adenoma",
    category_col = moran_column,
    condition_level = "spot",
    condition_keep = c("Normal", "Low-grade"),
    n_perm = 999,
    seed = 1,
    paired_only = TRUE
  )

  moran_adenoma_tests[[moran_column]] <- list(
    proportions = adenoma_test$tab,
    permanova = adenoma_test$permanova,
    betadisper = adenoma_test$betadisper_permutest,
    posthoc = adenoma_test$posthoc
  )

  adenoma_output_stem <- here(
    "results",
    "tables",
    paste0("local_moran_adenoma_", tolower(sub("^Cluster_Moran_", "", moran_column)))
  )
  write_moran_composition_outputs(adenoma_test, adenoma_output_stem)
}

saveRDS(
  moran_largesmall_tests,
  here("results", "tables", "local_moran_largesmall_tests.rds")
)
saveRDS(
  moran_adenoma_tests,
  here("results", "tables", "local_moran_adenoma_tests.rds")
)
saveRDS(
  list(largesmall = moran_largesmall_tests, adenoma = moran_adenoma_tests),
  here("results", "tables", "local_moran_composition_tests.rds")
)
```

# Partial spatial correlation

For each sample, the before-versus-after GDF15 comparison is performed on Fisher-transformed Spearman correlations. The exact paired Wilcoxon signed-rank test and the sign-flip permutation test therefore use `z = atanh(rho)`, while raw Spearman correlations and their difference are retained for interpretation and plotting. The size-stratified tests use the same exact Wilcoxon procedure; the Small-only result is considered exploratory because it contains five paired samples.

```{r partial-spatial-correlation}
lag_data <- spatial_metadata(
  combined_final,
  variables = c("GDF15", "Senepy", "Stem_Signature")
) |>
  filter(Adenoma == "Low-grade")

before_gdf15 <- within_sample_lagcorr(
  lag_data,
  x = "Senepy",
  y = "Stem_Signature",
  coords_cols = c("array_row", "array_col"),
  k = 6,
  n_perm = 999,
  seed = 1,
  mode = "x_vs_Wy"
)
after_gdf15 <- within_sample_lagcorr(
  lag_data,
  x = "Senepy",
  y = "Stem_Signature",
  coords_cols = c("array_row", "array_col"),
  k = 6,
  n_perm = 999,
  seed = 1,
  partial_covars = "GDF15",
  residualize_x = TRUE,
  mode = "x_vs_Wy"
)
partial_spatial_test <- compare_before_after(
  before_gdf15$per_sample,
  after_gdf15$per_sample,
  permutations = 999,
  seed = 1,
  label = "x_vs_Wy (smoothed)"
)
write_csv(partial_spatial_test$table, here("results", "tables", "partial_spatial_correlation.csv"))
write_csv(
  partial_spatial_test$stratified,
  here("results", "tables", "partial_spatial_correlation_stratified.csv")
)
write_csv(
  tibble(
    label = partial_spatial_test$label,
    n_pairs = nrow(partial_spatial_test$table),
    wilcox_statistic = unname(partial_spatial_test$paired_wilcox$statistic),
    wilcox_p = partial_spatial_test$paired_wilcox$p.value,
    paired_perm_p = partial_spatial_test$paired_perm_p,
    median_delta_z = partial_spatial_test$median_delta_z,
    median_rho_before = partial_spatial_test$median_rho_before,
    median_rho_after = partial_spatial_test$median_rho_after
  ),
  here("results", "tables", "partial_spatial_correlation_summary.csv")
)
saveRDS(
  partial_spatial_test,
  here("results", "tables", "partial_spatial_correlation_test.rds")
)

partial_spatial_test$paired_wilcox
partial_spatial_test$paired_perm_p
partial_spatial_test$stratified
```

# CD8 neighborhood analysis

The `ADT_dsb_by_slide` assay contains within-slide DSB-normalized values. GDF15 and the composite CD8 gate are standardized across all ADT-complete spots within each sample before restricting to normal and low-grade epithelial clusters C0, C1, C2, C3, C4, C6, and C8. CD8-positive spots require z-scored CD3E and CD8A of at least 0.5 and the maximum z-scored CD45RA/CD45RO value of at least zero. Epithelial tissue components use a 1.6-spot-unit connection radius, and CD8 density is measured within a two-spot-unit ring in the same component. The first comparison uses at least five GDF15-high source spots per pathology region; the second uses at least ten GDF15-high or GDF15-low source spots per region and excludes pilot samples.

```{r cd8-neighborhood}
cd8_data <- prepare_cd8_neighborhood_data(
  combined_final,
  adt_assay = "ADT_dsb_by_slide"
)
cd8_region <- cd8_density_by_source(
  cd8_data,
  comparison = "region",
  radius_multiplier = 2,
  min_sources = 5
)
cd8_gdf15 <- cd8_density_by_source(
  cd8_data,
  comparison = "gdf15",
  radius_multiplier = 2,
  min_sources = 10
)

cd8_region_test <- cd8_region |>
  dplyr::filter(!is.na(.data$dens_LGD), !is.na(.data$dens_Normal)) |>
  dplyr::group_by(.data$LargeSmall) |>
  dplyr::summarise(
    n_pairs = dplyr::n(),
    p_value = stats::wilcox.test(.data$dens_LGD, .data$dens_Normal, paired = TRUE)$p.value,
    median_difference = stats::median(.data$delta_dens),
    .groups = "drop"
  )
cd8_gdf15_test <- cd8_gdf15 |>
  dplyr::filter(!is.na(.data$dens_High), !is.na(.data$dens_Low)) |>
  dplyr::group_by(.data$Adenoma) |>
  dplyr::summarise(
    n_pairs = dplyr::n(),
    p_value = stats::wilcox.test(.data$dens_High, .data$dens_Low, paired = TRUE)$p.value,
    median_difference = stats::median(.data$delta_HL),
    .groups = "drop"
  )
write_csv(cd8_region, here("results", "tables", "cd8_density_region.csv"))
write_csv(cd8_gdf15, here("results", "tables", "cd8_density_gdf15_high_low.csv"))
write_csv(cd8_region_test, here("results", "tables", "cd8_density_region_tests.csv"))
write_csv(cd8_gdf15_test, here("results", "tables", "cd8_density_gdf15_high_low_tests.csv"))
```

# Adenoma radial distance from the tract surface

This section reconstructs `combined_final_semla` from `combined_final` and the Space Ranger coordinates, scale factors, and H&E images when `rebuild_staffli: true`. Manual annotations exported from the semla viewer use `All` for tract-surface anchors and `Island` for tissue pieces that were excluded. The distance is calculated within each semla-connected tissue component after removing `Island` annotations and singleton components. Surface anchors and spots within the first 100 micrometers are excluded from the summaries; remaining spots are grouped into 100-micrometer bins, and bins with fewer than five spots are removed.

```{r tract-surface-distance}
surface_annotations <- read_csv(
  here("data", "reference", "luminal_surface_annotations.csv"),
  show_col_types = FALSE
)
assert_columns(surface_annotations, c("sample_id", "barcode", "surface_label"))

if (isTRUE(params$rebuild_staffli)) {
  surface_object <- build_staffli_from_spaceranger(
    combined_final,
    manifest,
    image_height = 300
  )
  qs::qsave(
    surface_object,
    here("data", "derived", "visium_semla.qs"),
    preset = "high"
  )
} else {
  surface_object <- read_analysis_archive(
    here("data", "derived", "visium_semla.qs")
  )
}

surface_object <- separate_tissue_islands_semla(surface_object)
staffli_coordinates <- surface_object@tools$Staffli@meta_data |>
  as_tibble() |>
  dplyr::select(barcode, pxl_col_in_fullres, pxl_row_in_fullres)

surface_annotations <- surface_annotations |>
  transmute(
    sample_id = as.character(sample_id),
    raw_barcode = barcode,
    surface_label = case_when(
      toupper(surface_label) == "ALL" ~ "All",
      toupper(surface_label) == "ISLAND" ~ "Island",
      TRUE ~ as.character(surface_label)
    )
  )
surface_data <- surface_object@meta.data |>
  rownames_to_column("barcode") |>
  dplyr::select(barcode, Sample_ID, Adenoma, tissue_island, Senepy, Stem_Signature) |>
  mutate(raw_barcode = str_extract(barcode, "[ACGT]{16}-[0-9]+")) |>
  left_join(staffli_coordinates, by = "barcode") |>
  left_join(
    surface_annotations,
    by = c("Sample_ID" = "sample_id", "raw_barcode")
  ) |>
  mutate(
    surface_label = replace_na(surface_label, "Other"),
    Adenoma = recode(as.character(Adenoma), `Low-grade` = "Adenoma", `2` = "Normal")
  ) |>
  filter(
    surface_label != "Island",
    !tolower(tissue_island) %in% c("singleton", "singletons")
  )

surface_distance <- distance_to_surface(
  surface_data,
  surface_col = "surface_label",
  surface_value = "All"
)
surface_data <- left_join(
  surface_data,
  surface_distance,
  by = c("Sample_ID", "tissue_island", "barcode")
)

write_csv(surface_data, here("results", "tables", "surface_distance_spots.csv"))
surface_analysis_data <- filter(surface_data, Adenoma != "Stroma")
stem_distance_bins <- summarize_distance_bins(surface_analysis_data, score = "Stem_Signature")
senescence_distance_bins <- summarize_distance_bins(surface_analysis_data, score = "Senepy")
write_csv(stem_distance_bins, here("results", "tables", "surface_distance_stemness.csv"))
write_csv(senescence_distance_bins, here("results", "tables", "surface_distance_senescence.csv"))
```

# Continuous and categorical-proportion tests

All continuous-value comparisons use `compare_score_clusters_paired()`. Spot-level conditions are summarized within each sample and compared with a paired Wilcoxon signed-rank test; sample-level conditions are summarized per sample and compared with an unpaired Wilcoxon rank-sum test. The reported effect is the median for `level_a` minus the median for `level_b`.

The Senepy comparisons correspond to the sample-level cluster summary plotted in `Supple_plotting.Rmd`. Pseudotime is compared between paired Low-grade and Normal regions, and between Large and Small adenomas after restricting the latter analysis to Low-grade spots. Every continuous comparison uses per-sample medians and requires at least 30 spots in an analyzed sample/condition stratum. Benjamini-Hochberg-adjusted p-values are reported across the three Senepy cluster contrasts.

All categorical-proportion comparisons use `run_comp_tests()`. The function forms a complete category-proportion table for every sample/condition stratum, applies the Hellinger transformation before PERMANOVA and its dispersion check, and performs category-wise Wilcoxon tests with Benjamini-Hochberg correction. Pathology is a spot-level condition, so Normal and Low-grade regions are paired within samples that contain both. Adenoma post-hoc effects are Low-grade minus Normal. Adenoma size is a sample-level condition, so Small and Large samples are compared as independent groups. The same helper can be used for other categorical outcomes, including Moran's I classifications, by changing `category_col`.

```{r sample-level-statistics}
senepy_c2_vs_c3 <- compare_score_clusters_paired(
  combined_final@meta.data,
  score_col = "Senepy",
  sample_col = "Sample_ID",
  condition_col = "combined_cluster",
  level_a = "2",
  level_b = "3",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

senepy_c4_vs_c3 <- compare_score_clusters_paired(
  combined_final@meta.data,
  score_col = "Senepy",
  sample_col = "Sample_ID",
  condition_col = "combined_cluster",
  level_a = "4",
  level_b = "3",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

senepy_c4_vs_c2 <- compare_score_clusters_paired(
  combined_final@meta.data,
  score_col = "Senepy",
  sample_col = "Sample_ID",
  condition_col = "combined_cluster",
  level_a = "4",
  level_b = "2",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

epithelial_metadata <- combined_final@meta.data |>
  filter(Adenoma %in% c("Normal", "Low-grade"))
pseudotime_lowgrade_vs_normal <- compare_score_clusters_paired(
  epithelial_metadata,
  score_col = "dpt_pseudotime",
  sample_col = "Sample_ID",
  condition_col = "Adenoma",
  level_a = "Low-grade",
  level_b = "Normal",
  condition_level = "spot",
  agg = "median",
  min_spots = 30
)

low_grade_metadata <- combined_final@meta.data |>
  filter(Adenoma == "Low-grade")
pseudotime_large_vs_small_lowgrade <- compare_score_clusters_paired(
  low_grade_metadata,
  score_col = "dpt_pseudotime",
  sample_col = "Sample_ID",
  condition_col = "LargeSmall",
  level_a = "Large",
  level_b = "Small",
  condition_level = "sample",
  agg = "median",
  min_spots = 30
)

continuous_score_tests <- list(
  senepy_c2_vs_c3 = senepy_c2_vs_c3,
  senepy_c4_vs_c3 = senepy_c4_vs_c3,
  senepy_c4_vs_c2 = senepy_c4_vs_c2,
  pseudotime_lowgrade_vs_normal = pseudotime_lowgrade_vs_normal,
  pseudotime_large_vs_small_lowgrade = pseudotime_large_vs_small_lowgrade
)

senepy_primary_p_adjusted <- p.adjust(
  c(
    senepy_c2_vs_c3 = senepy_c2_vs_c3$test$p.value,
    senepy_c4_vs_c3 = senepy_c4_vs_c3$test$p.value,
    senepy_c4_vs_c2 = senepy_c4_vs_c2$test$p.value
  ),
  method = "BH"
)

continuous_test_summary <- tibble(
  comparison = names(continuous_score_tests),
  score = c("Senepy", "Senepy", "Senepy", "dpt_pseudotime", "dpt_pseudotime"),
  condition = c(
    "combined_cluster", "combined_cluster", "combined_cluster", "Adenoma", "LargeSmall"
  ),
  level_a = c("2", "4", "4", "Low-grade", "Large"),
  level_b = c("3", "3", "2", "Normal", "Small"),
  mode = vapply(continuous_score_tests, `[[`, character(1), "mode"),
  n_a = vapply(
    continuous_score_tests,
    function(x) if (is.null(x$n_a)) x$n_samples else x$n_a,
    numeric(1)
  ),
  n_b = vapply(
    continuous_score_tests,
    function(x) if (is.null(x$n_b)) x$n_samples else x$n_b,
    numeric(1)
  ),
  median_a = vapply(continuous_score_tests, function(x) x$ave_median[[1]], numeric(1)),
  median_b = vapply(continuous_score_tests, function(x) x$ave_median[[2]], numeric(1)),
  median_difference_a_minus_b = vapply(continuous_score_tests, `[[`, numeric(1), "diff_median"),
  statistic = vapply(continuous_score_tests, function(x) unname(x$test$statistic), numeric(1)),
  p_value = vapply(continuous_score_tests, function(x) x$test$p.value, numeric(1)),
  p_adjusted_bh = c(unname(senepy_primary_p_adjusted), NA_real_, NA_real_)
)

write_csv(continuous_test_summary, here("results", "tables", "continuous_score_tests.csv"))
saveRDS(continuous_score_tests, here("results", "tables", "continuous_score_tests.rds"))
continuous_test_summary

# Supplementary Figure 4J uses the same paired, sample-level median strategy
# for the two ADT signals. ADT-incomplete spots remain missing after the join.
adt_region_features <- c("ADT-CD8A", "ADT-CD3E")
adt_joined <- JoinLayers(combined_final, assay = "ADT_dsb_by_slide")
adt_values <- FetchData(
  adt_joined,
  vars = adt_region_features,
  assay = "ADT_dsb_by_slide",
  layer = "data"
) 
adt_region_metadata = AddMetaData(combined_final,adt_values)
adt_region_metadata = adt_region_metadata@meta.data|>
  filter(Adenoma %in% c("Normal", "Low-grade"))

adt_region_tests <- set_names(adt_region_features) |>
  map(function(feature) {
    compare_score_clusters_paired(
      adt_region_metadata,
      score_col = feature,
      sample_col = "Sample_ID",
      condition_col = "Adenoma",
      level_a = "Low-grade",
      level_b = "Normal",
      condition_level = "spot",
      agg = "median",
      min_spots = 30
    )
  })

adt_region_test_summary <- imap_dfr(adt_region_tests, function(result, feature) {
  tibble(
    feature = feature,
    n_paired_samples = result$n_samples,
    median_low_grade = result$ave_median[[1]],
    median_normal = result$ave_median[[2]],
    median_paired_difference = result$diff_median,
    statistic = unname(result$test$statistic),
    p_value = result$test$p.value
  )
})
write_csv(
  adt_region_test_summary,
  here("results", "tables", "adt_region_tests.csv")
)
saveRDS(
  adt_region_tests,
  here("results", "tables", "adt_region_tests.rds")
)
adt_region_test_summary

test_adenoma <- run_comp_tests(
  combined_final@meta.data,
  sample_col = "Sample_ID",
  condition_col = "Adenoma",
  category_col = "combined_cluster",
  condition_level = "spot",
  condition_keep = c("Normal", "Low-grade"),
  n_perm = 999,
  seed = 1,
  paired_only = TRUE
)

test_largesmall <- run_comp_tests(
  combined_final@meta.data,
  sample_col = "Sample_ID",
  condition_col = "LargeSmall",
  category_col = "combined_cluster",
  condition_level = "sample",
  condition_keep = c("Small", "Large"),
  n_perm = 999,
  seed = 1,
  paired_only = FALSE
)

write_composition_outputs <- function(result, comparison_name) {
  output_prefix <- here("results", "tables", paste0("cluster_composition_", comparison_name))
  write_csv(result$tab, paste0(output_prefix, "_proportions.csv"))
  write_csv(result$posthoc, paste0(output_prefix, "_posthoc.csv"))
  write_csv(
    rownames_to_column(as.data.frame(result$permanova), "term"),
    paste0(output_prefix, "_permanova.csv")
  )
  writeLines(
    capture.output(result$betadisper_permutest),
    paste0(output_prefix, "_betadisper.txt")
  )
  invisible(result)
}

write_composition_outputs(test_adenoma, "adenoma")
write_composition_outputs(test_largesmall, "largesmall")
saveRDS(
  list(adenoma = test_adenoma, largesmall = test_largesmall),
  here("results", "tables", "cluster_composition_tests.rds")
)

test_adenoma$permanova
test_adenoma$betadisper_permutest
test_adenoma$posthoc
test_largesmall$permanova
test_largesmall$betadisper_permutest
test_largesmall$posthoc
```

# Independent public single-cell validation

This analysis uses the separate Lau discovery and validation objects; these are not the Cell2location reference. The public discovery and validation objects are restricted to tubular/tubulovillous adenomas before this step. Original published embeddings and annotations are retained in the complete objects; epithelial subsets are re-normalized and re-clustered for the subpopulation analysis.

```{r single-cell-validation}
validation_objects <- list(
  discovery = qs::qread(here("data", "reference", "validation_discovery.qs")),
  validation = qs::qread(here("data", "reference", "validation_validation.qs"))
)

recluster_epithelium <- function(object, epithelial_cells, resolution = 0.4) {
  epithelial <- subset(object, cells = epithelial_cells)
  DefaultAssay(epithelial) <- "RNA"
  epithelial |>
    NormalizeData(verbose = FALSE) |>
    FindVariableFeatures(nfeatures = 2000, verbose = FALSE) |>
    ScaleData(verbose = FALSE) |>
    RunPCA() |>
    FindNeighbors() |>
    FindClusters(resolution = resolution, algorithm = 3, verbose = FALSE) |>
    RunUMAP(dims = 1:30, seed.use = 1, verbose = FALSE)
}

score_validation_object <- function(object) {
  cells <- rownames(object@meta.data)[
    !is.na(object$Cell_Type) & object$Cell_Type != "SSC"
  ]
  object <- subset(object, cells = cells)
  object <- run_irgsea(object, gene_sets, group_by = "Cell_Type", assay = "RNA")
  object <- copy_ucell_scores(
    object,
    c(
      Stem_Signature = "Stem_Signature",
      IEX = "IEX_score",
      WNT_BETA_CATENIN_SIGNALING = "HALLMARK_WNT_BETA_CATENIN_SIGNALING"
    )
  )
  object <- Calculate_senepy(object, assay = "RNA", layer = "counts",binarize = T)
  object$Senepy <- object$epithelial_cell_score_0
  object
}

validation_objects <- purrr::map(validation_objects, score_validation_object)
validation_epithelium <- purrr::imap(validation_objects, function(object, cohort) {
  epithelial_cells <- rownames(object@meta.data)[object$Cell_Type == "ASC"]
  recluster_epithelium(
    object,
    epithelial_cells,
    resolution = 0.4
  )
})
qs::qsave(
  validation_objects,
  here("data", "derived", "single_cell_validation_full.qs")
)

qs::qsave(validation_epithelium, here("data", "derived", "single_cell_validation.qs"))

validation_non_epi = qs::qread(here("data", "reference", "validation_non_epi.qs"))
validation_non_epi <- Calculate_senepy(validation_non_epi, assay = "RNA", layer = "counts",binarize = T)

qs::qsave(validation_non_epi, here("data", "derived", "single_cell_validation_non_epi.qs"))
```

```{r save-final-object}
qs::qsave(combined_final, here("data", "derived", "visium_scored.qs"), preset = "high")

# Cell2location abundances and local-Moran classifications are calculated after
# the diffusion map. Copy those later metadata columns into the epithelial
# object by barcode before writing the final plotting checkpoint.
epithelial_metadata_columns <- setdiff(
  colnames(combined_final@meta.data),
  colnames(epithelial@meta.data)
)
if (length(epithelial_metadata_columns)) {
  stopifnot(all(colnames(epithelial) %in% rownames(combined_final@meta.data)))
  epithelial <- AddMetaData(
    epithelial,
    combined_final@meta.data[
      colnames(epithelial),
      epithelial_metadata_columns,
      drop = FALSE
    ]
  )
}
qs::qsave(
  epithelial,
  here("data", "derived", "epithelial_diffusion_map.qs"),
  preset = "high"
)
writeLines(capture.output(sessionInfo()), here("results", "sessionInfo.txt"))
```
